---
title: "Getting started with hsquared"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with hsquared}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`hsquared` is the R-facing package for an open quantitative-genetic modelling
system. Its partner package, `HSquared.jl`, owns the sparse Julia engine.

The first user contract is intentionally simple:

```{r, eval = FALSE}
fit <- hsquared(
  y ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE
)
```

Current status: v0.1 **fits** this model. `hsquared()` parses and validates the
formula, marshals the model to the `HSquared.jl` engine, and returns a fitted
`hsquared_fit` object carrying heritability, variance components, breeding values
(EBVs/BLUPs), fixed effects, fitted values, residuals, and diagnostics, estimated
by REML (average-information). Fitting is computed in `HSquared.jl`, so it needs a
local Julia, the `JuliaCall` R package, and an `HSquared.jl` checkout; without
them the default call errors with install guidance, and
`control = hs_control(engine = "validate")` validates the contract without
fitting. ML is not implemented — `REML = FALSE` is rejected on the fit path. The
fit is validated by known-truth recovery, the published gryphon REML anchor, and
agreement with the `sommer` package. `hs_data()` can collect phenotype, pedigree,
genotype, marker, expression, annotation, and environment inputs into a checked
container for future integrated workflows.

The interface rule is: easy, easy, easy. Users are gold. The common animal
model should be obvious before users reach for genomic, multivariate,
factor-analytic, non-Gaussian, or unusual-inheritance machinery.

## Engine setup

Installing `hsquared` (`pak::pak("itchyshin/hsquared")`) gives you the R
interface; it does not install the engine. `HSquared.jl` is a from-source Julia
checkout, not a package-managed dependency. Install
[Julia](https://julialang.org/downloads/) and the `JuliaCall` R package, clone
the engine (`git clone https://github.com/itchyshin/HSquared.jl`), then register
the checkout in one of two ways:

```{r, eval = FALSE}
# (a) for the session, or persistently via .Renviron
Sys.setenv(HSQUARED_JULIA_PROJECT = "/path/to/HSquared.jl")

# (b) per call
fit <- hsquared(
  y ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  control = hs_control(
    engine_control = list(julia_project = "/path/to/HSquared.jl")
  )
)
```

Until the engine is registered, `control = hs_control(engine = "validate")`
parses and validates the model without fitting.

## Current parser contract

The parser currently accepts:

- one `animal()` term;
- random-intercept syntax: `1 | id`;
- `pedigree = ped`, or `animal(1 | id)` when `data` is an `hs_data()` bundle
  with a pedigree component;
- numeric Gaussian responses.

On this held candidate branch, one narrow experimental genomic GREML route is
available through the ordinary no-control call; the explicit
`engine = "julia", target = "genomic"` compatibility route remains supported:

```r
y ~ fixed + genomic(1 | id, markers = M)
y ~ fixed + genomic(1 | id, Ginv = Q)
```

Marker input uses sample-frequency, unweighted VanRaden method 1 with
`K_lambda = G + 0.01I`; supplied `Ginv` is used unchanged and its construction
details remain unknown. `heritability()` labels

`genomic_variance_ratio = sigma_g2 / (sigma_g2 + sigma_e2)`,

the genomic variance-component ratio on the declared relationship scale. It is
not generally the fraction of average marginal phenotypic variance and is not
pedigree-, founder-base-, population-, or universal narrow-sense heritability;
genomic interval and ratio-SE accessors are unavailable.
A first offset-5001 boundary candidate remains a banked 5.99x runtime-gate
failure. The separately preregistered revised offset-6001 candidate passed its
fresh holdout (240/240 valid, 40 wins, 0 losses, maximum p95 ratio 1.370),
clearing only the performance gate. The offset-7001 infrastructure attempt
failed before fitting because the bridge package was unavailable; it produced
no recovery summary or recovery evidence, and offsets 7001:7048 are retired.
The offset-7101 pilot completed 432/432 successful converged fits, but its
sealed adjudicator stopped on a logical serialization defect before minting a
receipt. All three summaries say `PRECISION_BLOCKER`, five cells exceed the
2,000-fit cap, no confirmation manifest exists, and offsets 7101:7148 are
retired. Recovery-v3 subsequently produced three separate 576-fit D0F corpora,
all permanently unadjudicated after exact replay stopped before row 1 for
fixed-panel cardinality, concrete-`Cmd` typing, and missing successful-gradient
contract failures. Their phenotype/bootstrap seeds are retired. Retry 4 is
repaired with disjoint seeds but is not presealed or run. This is diagnostic
rather than accepted recovery evidence; the route remains
partial/held and is not activated on main or released. `public_covered_count`
remains 5. The holdout used independent
HWE/no-LD markers. It provides no robustness
evidence for LD, population structure, imputation, base-frequency
misspecification, real panels, or production genotype data.

Several other terms fit through an opt-in experimental engine path
(`engine = "julia"`, not the default). The
common-environment leg of the two-effect model (`common_env()`) is **covered at
validation scale**; the remaining opt-in terms each mirror a `partial`
validation gate: `permanent()`, `maternal_genetic()`, genomic SNP-BLUP,
`single_step()`, and the supplied-relationship primaries `relmat()`
and `precision()` (REML-only, experimental, **not** covered — they do not change
the covered-model count). See the "Fitting models" article for the opt-in calls.

The remaining grammar is still important because it defines the roadmap. Terms
such as `cov = fa(K = 2)`, QTL effects, `paternal_genetic()`, `dominance()`,
`epistasis()`, selfing, polyploidy, haplodiploidy, and GLLVM-style latent
genetic axes are planned lanes, not current fitting features.

## Twin-package flow

The intended flow is:

```text
R formula + data
-> R validation and model specification
-> R bridge payload: y, X, sparse Z, method, IDs, pedigree metadata
-> HSquared.jl sparse engine
-> compact result
-> hsquared_fit object
-> R summaries, extractors, plots, and diagnostics
```

R owns the friendly interface. Julia owns speed, sparse computation, and
experimental engine work.
