---
title: "Great Tit Animal Model — ASReml (clutch size and laying date)"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    number-sections: true
    self-contained: true
execute:
  warning: false
  message: false
  cache: false
---

<!--
This notebook is the ASReml-only leg of the great tit comparison. Do not add
`library(JuliaCall)` or `library(hsquared)` here: asreml's bundled libomp.dylib
and Julia's OpenBLAS/MKL libomp.dylib both initialize LLVM's OpenMP runtime,
which crashes the R session with "OMP: Error #15" when loaded together. See
dev-test/great_tit_hsquared_model.qmd for the Julia-engine leg and
dev-test/great_tit_model_summary.qmd for the combined comparison — both are
rendered as separate R sessions and exchange results via saved .rds files.
-->

```{r}
#| label: packages
library(asreml)      # REML mixed-model engine with animal-model support
library(nadiv)       # pedigree sorting and founder completion (prepPed)
library(MCMCglmm)    # inverseA(): robust Ainv with inbreeding + logDet
library(Matrix)      # tril() / dgTMatrix for lower-triangle extraction
library(tidyverse)   # data wrangling and plotting
library(knitr)       # table rendering
```

# Data

## Load

```{r}
#| label: load-data
pheno <- read_csv(
  "dev-test/great_tit_breeding_data.csv",
  show_col_types = FALSE
)

# Pedigree has 8 columns; only the first three (id, dam, sire) are populated.
ped_raw <- read.csv(
  "dev-test/great_tit_pedigree.csv", header = TRUE, na.strings = NA)
```

## Inspect

```{r}
#| label: inspect-pheno
glimpse(pheno)
```

```{r}
#| label: inspect-ped
glimpse(ped_raw)
cat("Pedigree rows:", nrow(ped_raw), "\n")
cat("Unique IDs:   ", n_distinct(ped_raw$id), "\n")
```

```{r}
#| label: trait-summaries
pheno |>
  select(clutch_size, laydate) |>
  summary()
```

```{r}
#| label: trait-distributions
#| fig-width: 8
#| fig-height: 4
pheno |>
  select(clutch_size, laydate) |>
  pivot_longer(everything(), names_to = "trait", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  facet_wrap(~trait, scales = "free") +
  labs(x = NULL, y = "Count") +
  theme_bw()
```

# Pedigree

## Prepare with `nadiv`

`nadiv::prepPed()` sorts the pedigree so that parents always precede offspring,
adds any animals referenced as parents but absent as founders, and converts
`NA` unknowns to `0` (the convention ASReml expects is NA, but `prepPed` keeps
them as NA — `nadiv::makeAinv` handles both).

```{r}
#| label: prep-ped
# Replace "NA" strings with true NA in case any slipped through as character
ped_raw <- ped_raw |>
  mutate(across(everything(), ~ na_if(as.character(.), "NA")))

# prepPed requires a data.frame with columns id, dam, sire
ped <- prepPed(as.data.frame(ped_raw))

cat("Pedigree rows after prepPed:", nrow(ped), "\n")
```

## Build the inverse numerator relationship matrix

All attempts to coerce `nadiv`/`spam`/Matrix S4 objects into the format
`vm()` accepts failed. Instead we use `MCMCglmm::inverseA()` to obtain a
clean `dgCMatrix` plus inbreeding coefficients and a log-determinant, then
assemble these into an ASReml `ginv`-class object directly from its published
specification:

- 3-column numeric matrix: row index, column index, value (lower triangle,
  sorted row-wise)
- `class = "ginv"`
- attributes: `rowNames`, `inbreeding`, `geneticGroups`, `logdet`

```{r}
#| label: build-ainv
# Compute Ainv via MCMCglmm (handles pedigree prep internally)
Ainv <- ainverse(ped)
```

# Data preparation

## Link phenotypes to pedigree

The animal identifier in the phenotype file is `female_ID`.  We keep only
females whose ID appears in the pedigree.

```{r}
#| label: prep-pheno
dat <- pheno |>
  # Rename to 'animal' — the conventional name that ASReml looks for
  rename(animal = female_ID) |>
  # Keep records where the animal is in the pedigree
  filter(animal %in% ped$id) |>
  mutate(
    animal   = factor(animal),
    year_f   = factor(year),
    round_f  = factor(round),
    exp_manip = factor(experimental_manipulation)
  )

cat("Records after pedigree filter:", nrow(dat), "\n")
cat("Unique females:               ", n_distinct(dat$animal), "\n")
```

```{r}
#| label: missing-traits
dat |>
  summarise(
    n_clutch_size = sum(!is.na(clutch_size)),
    n_laydate     = sum(!is.na(laydate))
  ) |>
  kable(caption = "Records with non-missing trait values")
```

## Prepare per-trait analysis frames

We drop rows with `NA` in the response before passing to ASReml, because
ASReml can handle missing data but it is cleaner to be explicit here.

```{r}
#| label: dat-cs
dat_cs <- dat |>
  filter(!is.na(clutch_size)) |>
  droplevels()

dat_ld <- dat |>
  filter(!is.na(laydate)) |>
  droplevels()

dat_cs$animal2 <- dat_cs$animal
dat_ld$animal2 <- dat_ld$animal
```

# Animal model — clutch size

## Model specification

Fixed effects: `year_f` (accounts for annual mean differences), `round_f`
(first vs replacement clutch), `exp_manip` (experimental manipulation flag),
`female.age` (linear age effect).

Random effects: `vm(animal, Ainv)` — the additive genetic effect fitted
against the pedigree-derived inverse relationship matrix built by
`asreml::ainverse()`. `ide(animal)` is the permanent environmental effect
for repeated measures (a female can appear in multiple years).

```{r}
#| label: fit-cs
#| results: hide
asreml.options(workspace = "1gb", maxit = 50, trace = TRUE)

time_cs <- system.time({
  fit_cs <- asreml(
    fixed    = clutch_size ~ year_f + round_f + exp_manip + female.age,
    random   = ~ vm(animal, Ainv) + ide(animal),
    data     = dat_cs,
    na.action = na.method(x = "include", y = "omit")
  )
  # fit_cs <- update(fit_cs)
})
```

```{r}
#| label: time-cs
cat(sprintf("ASReml clutch size  — elapsed: %.1f s  (user: %.1f s, sys: %.1f s)\n",
            time_cs["elapsed"], time_cs["user.self"], time_cs["sys.self"]))
```

## Convergence check

```{r}
#| label: convergence-cs
cat("Converged:", fit_cs$converge, "\n")
summary(fit_cs)$varcomp |>
  kable(digits = 4, caption = "Variance components — clutch size")
```

## Heritability

$$
h^2 = \frac{\sigma^2_A}{\sigma^2_A + \sigma^2_{PE} + \sigma^2_e}
$$

`vpredict()` computes the estimate and its standard error via the delta method.

```{r}
#| label: h2-cs
vc_cs <- summary(fit_cs)$varcomp

# ASReml component names: vm(animal, Ainv)!vm, ide(animal)!ide, units!R
h2_cs <- vpredict(
  fit_cs,
  h2 ~ V1 / (V1 + V2 + V3)
)

cat("h² clutch size:", round(h2_cs[["Estimate"]], 3),
    "±", round(h2_cs[["SE"]], 3), "(SE)\n")
```

## Wald statistics for fixed effects

```{r}
#| label: wald-cs
wald(fit_cs, denDF = "numeric", ssType = "conditional") |>
  kable(digits = 3, caption = "Wald F-statistics — clutch size")
```

# Animal model — laying date

## Model specification

Same fixed and random structure as the clutch size model.

```{r}
#| label: fit-ld
#| results: hide
time_ld <- system.time({
  fit_ld <- asreml(
    fixed    = laydate ~ year_f + round_f + exp_manip + female.age,
    random   = ~ vm(animal, Ainv) + ide(animal),
    data     = dat_ld,
    na.action = na.method(x = "include", y = "omit")
  )
  fit_ld <- update(fit_ld)
})
```

```{r}
#| label: time-ld
cat(sprintf("ASReml laying date  — elapsed: %.1f s  (user: %.1f s, sys: %.1f s)\n",
            time_ld["elapsed"], time_ld["user.self"], time_ld["sys.self"]))
```

## Convergence check

```{r}
#| label: convergence-ld
cat("Converged:", fit_ld$converge, "\n")
summary(fit_ld)$varcomp |>
  kable(digits = 4, caption = "Variance components — laying date")
```

## Heritability

```{r}
#| label: h2-ld
h2_ld <- vpredict(
  fit_ld,
  h2 ~ V1 / (V1 + V2 + V3)
)

cat("h² laying date:", round(h2_ld[["Estimate"]], 3),
    "±", round(h2_ld[["SE"]], 3), "(SE)\n")
```

## Wald statistics for fixed effects

```{r}
#| label: wald-ld
wald(fit_ld, denDF = "numeric", ssType = "conditional") |>
  kable(digits = 3, caption = "Wald F-statistics — laying date")
```

# Summary

```{r}
#| label: summary-table
asreml_vc_table <- function(fit) {
  summary(fit)$varcomp |>
    as.data.frame() |>
    tibble::rownames_to_column("Component") |>
    transmute(Component, Estimate = component, SE = std.error)
}

vc_cs_df <- asreml_vc_table(fit_cs)
vc_ld_df <- asreml_vc_table(fit_ld)

tibble(
  Trait            = c("Clutch size", "Laying date"),
  `σ²_A`           = c(vc_cs_df$Estimate[vc_cs_df$Component == "vm(animal, Ainv)"],
                        vc_ld_df$Estimate[vc_ld_df$Component == "vm(animal, Ainv)"]),
  `σ²_PE`          = c(vc_cs_df$Estimate[vc_cs_df$Component == "ide(animal)"],
                        vc_ld_df$Estimate[vc_ld_df$Component == "ide(animal)"]),
  `σ²_e`           = c(vc_cs_df$Estimate[vc_cs_df$Component == "units!R"],
                        vc_ld_df$Estimate[vc_ld_df$Component == "units!R"]),
  `h² (estimate)`  = c(h2_cs[["Estimate"]], h2_ld[["Estimate"]]),
  `h² (SE)`        = c(h2_cs[["SE"]],       h2_ld[["SE"]])
) |>
  mutate(across(where(is.numeric), ~ round(., 3))) |>
  kable(caption = "Heritability summary — great tit Wytham Woods population")
```

## Save results for the cross-engine summary

Persist everything the `hsquared` leg and the combined summary notebook need,
so neither has to load `asreml` (and therefore never loads in the same R
session as `JuliaCall`).

```{r}
#| label: save-results
saveRDS(
  list(
    dat_cs   = dat_cs,
    dat_ld   = dat_ld,
    ped      = ped,
    vc_cs    = vc_cs_df,
    vc_ld    = vc_ld_df,
    h2_cs    = h2_cs,
    h2_ld    = h2_ld,
    time_cs  = time_cs,
    time_ld  = time_ld
  ),
  file = "dev-test/asreml_results.rds"
)
```

# Session info

```{r}
#| label: session
sessionInfo()
```
