---
title: "Great Tit Animal Model — hsquared (Julia engine)"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    number-sections: true
    self-contained: true
execute:
  warning: false
  message: false
  cache: false
---

<!--
This notebook is the hsquared/Julia-engine leg of the great tit comparison.
Do not add `library(asreml)` here: asreml's bundled libomp.dylib and Julia's
OpenBLAS/MKL libomp.dylib both initialize LLVM's OpenMP runtime, which crashes
the R session with "OMP: Error #15" when loaded together. Render
dev-test/great_tit_asreml_model.qmd first (it saves
dev-test/asreml_results.rds), then this notebook, then
dev-test/great_tit_model_summary.qmd for the combined comparison.
-->

```{r}
#| label: packages
library(tidyverse)   # data wrangling
library(knitr)       # table rendering
library(devtools)    # load_all() for hsquared development checkout
library(JuliaCall)

# remotes::install_github("itchyshin/hsquared")
```

## Setup

```{r}
#| label: hs-setup
# Point the bridge at the local HSquared.jl checkout.
# The engine reads this before JuliaCall starts.
Sys.setenv(
  HSQUARED_JULIA_PROJECT =
    "/Users/szymek/Library/CloudStorage/OneDrive-UniwersytetJagielloński/Science/Julia/HSquared.jl"
)

Sys.setenv(
  PATH = paste(
    "/Users/szymek/.juliaup/bin",
    Sys.getenv("PATH"),
    sep = .Platform$path.sep
  )
)

# Load hsquared from the local development checkout
# devtools::load_all()
library(hsquared)
```

## Load the ASReml leg's saved data and pedigree

```{r}
#| label: load-asreml-results
asreml_results <- readRDS("dev-test/asreml_results.rds")
dat_cs <- asreml_results$dat_cs
dat_ld <- asreml_results$dat_ld
ped    <- asreml_results$ped
```

## Bundle phenotypes and pedigree

`hs_data()` pairs the phenotype table with the prepared pedigree. The pedigree
must have columns `id`, `dam`, `sire` — exactly the output of `prepPed()`.

The full pedigree includes many individuals unrelated to the females retained
for these two analyses. Keep those females and every known ancestor before
passing the data to the Julia engine.

```{r}
#| label: hs-bundle
analysis_ids <- unique(c(as.character(dat_cs$animal), as.character(dat_ld$animal)))
pedigree_ids <- analysis_ids

repeat {
  pedigree_rows <- match(pedigree_ids, ped$id)
  parent_ids <- unique(c(ped$dam[pedigree_rows], ped$sire[pedigree_rows]))
  expanded_ids <- unique(c(pedigree_ids, parent_ids[!is.na(parent_ids)]))

  if (length(expanded_ids) == length(pedigree_ids)) {
    break
  }
  pedigree_ids <- expanded_ids
}

ped_analysis <- ped[ped$id %in% pedigree_ids, c("id", "dam", "sire")]
stopifnot(all(analysis_ids %in% ped_analysis$id))

cat("Pedigree rows for hsquared:", nrow(ped_analysis), "\n")

bundle_cs <- hs_data(phenotypes = dat_cs, pedigree = ped_analysis, id = "animal")
bundle_ld <- hs_data(phenotypes = dat_ld, pedigree = ped_analysis, id = "animal")

data_status(bundle_cs)
```

## Animal model — clutch size

```{r}
#| label: hs-fit-cs
#| results: hide
time_hs_cs <- system.time({
    hs_fit_cs <- hsquared(
        # MATCHES THE ASReml MODEL: `~ vm(animal, Ainv) + ide(animal)` is the
        # additive genetic effect PLUS a permanent-environment effect. Fitting
        # `animal(1 | animal)` alone here made the comparison structurally
        # mismatched -- with 11,856 records on 7,340 females, the additive
        # variance absorbed the permanent-environment variance and came back as
        # a single 1.110 against ASReml's 0.595 + 0.525.
        #
        # `scale_method = "auto"` is required, not optional: the dense
        # estimator is capped at nobs^2 + nanimals^2 <= max_dense_cells, and
        # this problem needs ~2.6e8 cells against a 1e6 cap.
        clutch_size ~ year_f + round_f + exp_manip + female.age +
            animal(1 | animal) + permanent(1 | animal),
        data = bundle_cs,
        family = gaussian(),
        control = hs_control(
            engine = "julia",
            engine_control = list(
                target = "repeatability",
                scale_method = "auto"
            )
        ),
        REML = TRUE
    )
})
```

```{r}
#| label: time-hs-cs
cat(sprintf("hsquared clutch size — elapsed: %.1f s  (user: %.1f s, sys: %.1f s)\n",
            time_hs_cs["elapsed"], time_hs_cs["user.self"], time_hs_cs["sys.self"]))
```

```{r}
#| label: hs-vc-cs
variance_components(hs_fit_cs) |>
  kable(digits = 4, caption = "Variance components — clutch size (hsquared)")

heritability(hs_fit_cs) |>
  kable(digits = 4, caption = "Heritability — clutch size (hsquared)")
```

## Animal model — laying date

```{r}
#| label: hs-fit-ld
#| results: hide
time_hs_ld <- system.time({
  hs_fit_ld <- hsquared(
    laydate ~ year_f + round_f + exp_manip + female.age +
      animal(1 | animal) + permanent(1 | animal),
    data   = bundle_ld,
    family = gaussian(),
    control = hs_control(
      engine = "julia",
      engine_control = list(target = "repeatability", scale_method = "auto")
    ),
    REML   = TRUE
  )
})
```

```{r}
#| label: time-hs-ld
cat(sprintf("hsquared laying date — elapsed: %.1f s  (user: %.1f s, sys: %.1f s)\n",
            time_hs_ld["elapsed"], time_hs_ld["user.self"], time_hs_ld["sys.self"]))
```

```{r}
#| label: hs-vc-ld
variance_components(hs_fit_ld) |>
  kable(digits = 4, caption = "Variance components — laying date (hsquared)")

heritability(hs_fit_ld) |>
  kable(digits = 4, caption = "Heritability — laying date (hsquared)")
```

## Save results for the cross-engine summary

```{r}
#| label: save-results
hsquared_vc_table <- function(fit) {
  # hsquared DOES return asymptotic variance-component standard errors on this
  # route -- an earlier report hardcoded `SE = NA_real_` and never asked for
  # them, which is where the "hsquared produces no SEs" claim came from.
  se <- variance_component_standard_errors(fit)
  variance_components(fit) |>
    left_join(se, by = "component") |>
    transmute(Component = component, Estimate = estimate, SE = se)
}

hs_vc_cs <- hsquared_vc_table(hs_fit_cs)
hs_vc_ld <- hsquared_vc_table(hs_fit_ld)

hs_h2_cs <- heritability(hs_fit_cs) |>
  transmute(Estimate = estimate,
            SE = as.numeric(heritability_standard_error(hs_fit_cs)))
hs_h2_ld <- heritability(hs_fit_ld) |>
  transmute(Estimate = estimate,
            SE = as.numeric(heritability_standard_error(hs_fit_ld)))

saveRDS(
  list(
    vc_cs   = hs_vc_cs,
    vc_ld   = hs_vc_ld,
    h2_cs   = hs_h2_cs,
    h2_ld   = hs_h2_ld,
    time_cs = time_hs_cs,
    time_ld = time_hs_ld
  ),
  file = "dev-test/hsquared_results.rds"
)
```

# Session info

```{r}
#| label: session
sessionInfo()
```
