---
title: "Can I fit and report this?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Can I fit and report this?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This page answers one question: **can I use `hsquared` for my analysis, and what
am I allowed to write in the paper?**

`hsquared` is an experimental package. Most of what follows is not about whether
code runs — nearly all of it runs. It is about which numbers have enough evidence
behind them to report.

## The three-line version

1. Three **default-routed covered** paths exist: the **univariate Gaussian animal
   model**, **t = 2 unstructured multivariate** `cbind()` (validation-scale), and
   **genomic GREML** via narrow `genomic(...)` (validation-scale; experimental
   0.7.0 retained). Report their point estimates within the stated scope.
2. Almost everything else fits through an **opt-in** engine target (some covered
   at validation scale, some experimental). k≥3 and diagonal multivariate stay
   experimental. Treat experimental fits as exploratory, or report them beside a
   comparator you ran yourself.
3. **No standard error or interval in this package is coverage-calibrated.** Do
   not report one as if it were.

```{r f3-recovery-ladder, echo = FALSE, results = "asis"}
cat(sprintf(
  '<p><img src="../reference/figures/recovery-ladder.svg" class="r-plt" alt="%s" width="100%%"></p>\n',
  "Three-rung validation ladder from planned to partial to covered. A side note marks internal self-agreement as insufficient for covered status."
))
```

<p class="hs-banner" role="note"><strong>Covered at validation scale · experimental · dense · REML-only</strong> still means experimental packaging. It is not 1.0 and not interval-calibrated. The ladder caption is dated to this source tree; the live count is <code>validation_status()</code> and this limits page after any covered-row update — this figure does not flip a row. <code>public_covered_count</code> stays 7.</p>

## What the status words mean

`hsquared` uses four words, in descending order of what you may claim. It
deliberately never uses a bare "supported", because that word hides the
difference between *runs* and *validated*.

| Word | The code | The evidence | What you may report |
|---|---|---|---|
| `covered` | Runs | Pre-declared recovery gate passed **and** an external same-estimand comparator agrees | Point estimates, within the stated scope |
| `partial` | Runs | Bridge-verified and partly evidenced; the chain has a named hole | Exploratory use; report only beside your own comparator |
| `experimental` | Runs | Labelling, not a status tier — every opt-in route carries it | Nothing on this package's word alone |
| `planned` | Aborts as planned, not implemented | None | Nothing |

This page is the user-facing list of what you may report.
`validation_status()` is a developer evidence table of validation atoms;
it is not a complete list of covered routes. Developers can inspect it
in their own session:

```{r validation-status-live, eval = FALSE}
validation_status()
```

## Do I need Julia?

Yes, for fitting. This matters more here than in most R packages, because the
default call is the one that needs it.

| What you want | What you need |
|---|---|
| Parse and validate a model, inspect the design matrices and pedigree ordering | R only — `hs_control(engine = "validate")` and `model_spec()` |
| Actually fit and get heritability, variance components, EBVs | A local Julia, the `JuliaCall` package, and an `HSquared.jl` checkout |

Without the engine, `hsquared()` still tells you whether your model is
expressible, and errors with install guidance if you ask it to fit. See the
engine setup section of the [README](https://github.com/itchyshin/hsquared#engine-setup).

```{r validate-only, eval = FALSE}
# no Julia required; same ped/dat as the README
hsquared(
  weight ~ sex + animal(1 | id, pedigree = ped),
  data = dat,
  control = hs_control(engine = "validate")
)

model_spec(
  weight ~ sex + animal(1 | id, pedigree = ped),
  data = dat
)
```

## Route-by-route reporting permissions

```{r capability-ledger-summary, echo = FALSE, results = "asis"}
cat(readLines("includes/capability-ledger-summary.md", warn = FALSE), sep = "\n")
```

## Two covered routes without their own status row

`validation_status()` does not yet carry a separate row for two routes that
`docs/design/capability-status.md` records as covered at validation scale:

- **Random regression / reaction norm, k = 2** —
  `animal(rr(age, order = 2) | id, pedigree = ped)` via
  `engine_control = list(target = "random_regression")`. A pre-declared 48-seed
  recovery gate passed and a `sommer` `leg()` same-estimand comparator agrees on
  all `K_g` entries and the residual. **k = 2 only.** The residual is homogeneous
  and there is no permanent-environment term, so `rr_heritability()` can
  *overstate* h²(t) for repeated-records designs. `K_g` values are not comparable
  across basis-normalization conventions.
- **Direct–maternal correlated 2×2 G** —
  `animal(1 | id, pedigree = ped) + maternal_genetic(1 | dam)` via
  `engine_control = list(target = "direct_maternal")`. A pre-declared 48-seed gate
  passed and a `sommer` `covm()` same-estimand comparator agrees. `heritability()`
  returns the labelled Willham triple — direct h²_d, maternal m², total h²_T, and
  r_am — never a bare scalar. Direct h² is **not** the total heritability, and a
  negative r_am is real and expected. Single relationship A only.

Both are opt-in, experimental, REML-only, and dense (n ≲ 1000). They are
listed here because this page — not `validation_status()` — is what you
should read before reporting a number. The cards above are generated from
the evidence table and will not mention them until those rows exist.

## Where to read next

- [Model status](model-status.html) — the same ledger organised by model family.
- [Validation evidence](validation-evidence.html) — the evidence chain behind each
  row, atom by atom.
- [Function map and cheat sheet](function-map-cheatsheet.html) — which function to
  call at which step.
- [Formula grammar roadmap](formula-grammar.html) — what the parser reserves for
  later, and what it refuses today.
