---
title: "Can I fit and report this?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Can I fit and report this?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This page answers one question: **can I use `hsquared` for my analysis, and what
am I allowed to write in the paper?**

`hsquared` is an experimental package. Most of what follows is not about whether
code runs; nearly all of it runs. It is about which numbers have enough evidence
behind them to report.

## The three-line version

1. Two **default-routed covered** paths exist: the **univariate Gaussian animal
   model** and **t = 2 unstructured multivariate** `cbind()` (validation-scale).
   **Genomic GREML** is also covered at validation scale, but only through its
   explicit `engine = "julia", target = "genomic"` route; ordinary-call /
   default-route genomic activation remains held (design-44 G5;
   `BOUNDARY_HOLDOUT_FAIL`). Report point estimates only within each stated
   scope.
2. Almost everything else fits through an **opt-in** engine target (some covered
   at validation scale, some experimental). k≥3 and diagonal multivariate stay
   experimental. Treat experimental fits as exploratory, or report them beside a
   comparator you ran yourself.
3. **No standard error or interval in this package is nominally
   coverage-calibrated.** Named univariate pedigree intervals have a narrower
   `directional-conservative` claim level, not a nominal-calibration claim. Do
   not report any interval as if nominal coverage had been established.
   The information matrix also differs by route: see
   [Which information matrix?](#which-information-matrix).

```{r f3-recovery-ladder, echo = FALSE, results = "asis"}
cat(sprintf(
  '<p><img src="../reference/figures/recovery-ladder.svg" class="r-plt" alt="%s" width="100%%"></p>\n',
  "Three-rung validation ladder from planned to partial to covered. A side note marks internal self-agreement as insufficient for covered status."
))
```

<p class="hs-banner" role="note"><strong>Covered at validation scale · experimental · dense · REML-only</strong> still means experimental packaging. It is not 1.0 and not interval-calibrated. The ladder caption is dated to this source tree; the live count is <code>validation_status()</code> and this limits page after any covered-row update. This figure does not flip a row. <code>public_covered_count</code> stays 7.</p>

## What the status words mean

`hsquared` uses four words, in descending order of what you may claim. It
deliberately never uses a bare "supported", because that word hides the
difference between *runs* and *validated*.

| Word | The code | The evidence | What you may report |
|---|---|---|---|
| `covered` | Runs | Pre-declared recovery gate passed **and** an external same-estimand comparator agrees | Point estimates, within the stated scope |
| `partial` | Runs | Bridge-verified and partly evidenced; the chain has a named hole | Exploratory use; report only beside your own comparator |
| `experimental` | Runs | Labelling, not a status tier; every opt-in route carries it | Nothing on this package's word alone |
| `planned` | Aborts as planned, not implemented | None | Nothing |

This page is the user-facing list of what you may report.
`validation_status()` is a developer evidence table of validation atoms;
it is not a complete list of covered routes. Developers can inspect it
in their own session:

```{r validation-status-live, eval = FALSE}
validation_status()
```

## Do I need Julia?

Yes, for fitting. This matters more here than in most R packages, because the
default call is the one that needs it.

| What you want | What you need |
|---|---|
| Parse and validate a model, inspect the design matrices and pedigree ordering | R only: `hs_control(engine = "validate")` and `model_spec()` |
| Actually fit and get heritability, variance components, EBVs | A local Julia, the `JuliaCall` package, and an `HSquared.jl` checkout |

Without the engine, `hsquared()` still tells you whether your model is
expressible, and errors with install guidance if you ask it to fit. See the
[Installation](https://github.com/itchyshin/hsquared#installation) section of
the README.

```{r validate-only, eval = FALSE}
# no Julia required; same ped/dat as the README
hsquared(
  weight ~ sex + animal(1 | id, pedigree = ped),
  data = dat,
  control = hs_control(engine = "validate")
)

model_spec(
  weight ~ sex + animal(1 | id, pedigree = ped),
  data = dat
)
```

## Route-by-route reporting permissions

```{r capability-ledger-summary, echo = FALSE, results = "asis"}
cat(readLines("includes/capability-ledger-summary.md", warn = FALSE), sep = "\n")
```

## Two covered routes without their own status row

`validation_status()` does not yet carry a separate row for two routes that
[`docs/design/capability-status.md`](https://github.com/itchyshin/hsquared/blob/main/docs/design/capability-status.md) records as covered at validation scale:

- **Random regression / reaction norm, k = 2:**
  `animal(rr(age, order = 2) | id, pedigree = ped)` via
  `engine_control = list(target = "random_regression")`. A pre-declared 48-seed
  recovery gate passed and a `sommer` `leg()` same-estimand comparator agrees on
  all `K_g` entries and the residual. **k = 2 only.** The residual is homogeneous
  and there is no permanent-environment term, so `rr_heritability()` can
  *overstate* h²(t) for repeated-records designs. `K_g` values are not comparable
  across basis-normalization conventions.
- **Direct–maternal correlated 2×2 G:**
  `animal(1 | id, pedigree = ped) + maternal_genetic(1 | dam)` via
  `engine_control = list(target = "direct_maternal")`. A pre-declared 48-seed gate
  passed and a `sommer` `covm()` same-estimand comparator agrees. `heritability()`
  returns the labelled Willham triple: direct h²_d, maternal m², total h²_T, and
  r_am; never a bare scalar. Direct h² is **not** the total heritability, and a
  negative r_am is real and expected. Single relationship A only.

Both are opt-in, experimental, and REML-only. The engine-enforced dense
rule on the default `animal()` route and `target = "repeatability"` is
`nobs^2 + nanimals^2 <= 1e6` (default `max_dense_cells`), counting every
animal in the pedigree, not just records. Unlike those guarded routes,
neither random regression nor direct-maternal currently accepts
`engine_control$max_dense_cells`, so they have no engine-enforced cell
cap. Read this page before reporting a number;
`validation_status()` is a developer evidence table.
The cards above are generated from
the evidence table and will not mention them until those rows exist.

## New partial FA and genetic GLLVM routes

The four-trait, one-factor Gaussian pedigree FA fit is available through an
explicit Julia expert-control target. It returns a fitted genetic covariance,
uniqueness estimates, genetic correlations, and diagnostics. The rank does not
by itself identify every loading and uniqueness configuration. The tested
fixture required a higher iteration limit. In a 200-seed Gaussian pedigree
study with four traits and three records per animal, 141 fits converged and
110 met the declared covariance-recovery diagnostics, using two ordinary
starts with 5000 iterations each. All 200 attempts were counted. Inspect
convergence, start disagreement and floor diagnostics before interpreting a
fit. See [Multivariate Gaussian animal models](multivariate.html) for the exact
call and the study's limits.

The three-trait, two-factor Poisson-log pedigree genetic GLLVM fit is also an
explicit expert-control route. It returns genetic covariance and correlations
plus trait-level conditional genetic modes on the log-rate scale. See
[Genetic GLLVM opt-in](genetic-gllvm.html) for the exact call and limits.

Both routes are **partial**. Use their point estimates for exploration; for a
paper, run a suitable comparator yourself. No interval, automatic factor rank,
missing-response fit, or broader family claim follows from these cells.

## Which information matrix?

SE and interval extractors do not all use the same information matrix.
Swapping one matrix for the other changes the SE (the two are not the
same estimand). The method itself is unchanged; this table only names
what each route already computes.

| Route | Information matrix | How it is built | Ratio SE / interval |
| --- | --- | --- | --- |
| Univariate animal model (default) | Average information | Analytic AI matrix from the REML fit | Delta method; logit scale for `heritability_interval()` |
| Repeatability / permanent environment | Observed information | Central finite-difference Hessian, `fd_step = 1e-4` | Logit-scale delta method for `repeatability_interval()` |
| Two-effect (`common_env` / maternal) | Observed information | Central finite-difference Hessian, `fd_step = 1e-4` | Logit-scale delta method for `heritability_interval()` and `common_env_proportion_interval()` |
| Multivariate `cbind()` | Observed information | Central finite-difference Hessian, `fd_step = 1e-4` | Delta method for `covariance_standard_errors()`; Fisher-z for genetic-correlation endpoints |

See `?variance_component_standard_errors`, `?heritability_interval`,
`?repeatability_interval`, `?common_env_proportion_interval`, and
`?covariance_standard_errors`. None of these intervals is nominally
coverage-calibrated.

## Where to read next

- [Fitting models](fitting-models.html): the smallest honest workflow after you
  choose a route.
- [Visualizing models](visualizing-models.html): diagnose a fitted result before
  extracting it.
- [Model status](model-status.html): the same ledger organised by model family.
- [Validation evidence](validation-evidence.html): the evidence chain behind each
  row, atom by atom.
- [Function map and cheat sheet](function-map-cheatsheet.html): which function to
  call at which step.
- [Formula grammar roadmap](formula-grammar.html): what the parser reserves for
  later, and what it refuses today.
- [Progression & evidence](progression-evidence.html): dated milestones that do
  not themselves change a reporting permission.
