---
title: "Fitting quantitative-genetic models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting quantitative-genetic models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

<p class="hs-banner" role="note"><strong>Default path vs opt-in targets.</strong>
<code>engine = "fit"</code> is the v0.1 univariate Gaussian REML path. Other models below use
<code>engine = "julia"</code> with an explicit <code>target</code> (experimental / opt-in), except
multivariate <code>cbind()</code>, which routes on the default call but keeps the experimental
label. See <a href="current-limits.html">Can I fit and report this?</a>.</p>

This article shows the models `hsquared` can fit. Fitting is computed in the
sibling `HSquared.jl` engine, so every example needs a local Julia, the
`JuliaCall` package, and an `HSquared.jl` checkout. Register the checkout with
`Sys.setenv(HSQUARED_JULIA_PROJECT = "/path/to/HSquared.jl")` or per call via
`control = hs_control(engine_control = list(julia_project = "/path/to/HSquared.jl"))`;
see the README or "Getting started" article for full setup. The code chunks are
not run when this page is built (the build host has no Julia).

All fits are by REML. ML is not implemented. `REML = FALSE` is rejected; use
`REML = TRUE` (the default). That is the same rule for the default fit and for
`engine = "validate"`.

```{r twin-bridge, echo = FALSE, results = "asis", eval = TRUE}
cat(sprintf(
  '<p><img src="../reference/figures/twin-bridge.svg" class="r-plt" alt="%s" width="100%%"></p>\n',
  "Two-column diagram of the R to Julia twin bridge. Formula language stays in R; y, X, Z CSC, pedigree indices, and target metadata cross the bridge; solvers stay in Julia; result_payload returns variance components, heritability, and breeding values. Default engine equals fit for v0.1; other targets are opt-in experimental."
))
```

<p class="hs-caption"><em>What crosses the bridge vs what stays lane-local. Default
<code>engine = "fit"</code> is the v0.1 path; opt-in targets need
<code>engine = "julia"</code> and an explicit <code>target</code>. This figure does not imply
production bridge coverage for every model.</em></p>

## The default: a univariate Gaussian animal model

This is the v0.1 model, fitted by the default `hsquared()` call. It estimates the
additive-genetic and residual variances, heritability, and breeding values.

```{r}
library(hsquared)

fit <- hsquared(
  weight ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE
)

variance_components(fit)
heritability(fit)
breeding_values(fit)   # EBVs (aliases: EBV(), BLUP())
summary(fit)
```

Most of the remaining models below are **opt-in**: they are reached through
`engine = "julia"` with an explicit `target`. The one exception is the
multivariate `cbind()` model, which routes on the default call — being reachable
by default does not make it any less experimental. They are all REML-only,
experimental, and none is a covered public default. Four of them are
**covered at validation scale** —
each passed a pre-declared 48-seed recovery gate plus a same-estimand REML
comparator: the common-environment two-effect model, arbitrary-N multi-effect,
random regression k = 2, and the direct–maternal 2×2 G model. Covered at
validation scale means experimental and not production; none is a default fit.
The remaining opt-in models (`validation_status()` rows `partial`) have not yet
passed a comparator- or known-truth-validated gate.

## Repeatability (permanent environment)

A repeated-records model with an additive-genetic effect and a
permanent-environment effect that shares the animal grouping. The additive
(`σ²a`) and permanent-environment (`σ²pe`) variances are only identifiable with
repeated records per individual.

```{r}
fit_rep <- hsquared(
  y ~ animal(1 | id, pedigree = ped) + permanent(1 | id),
  data = repeated_dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "repeatability")
  )
)

variance_components(fit_rep)    # animal, permanent, residual
repeatability(fit_rep)          # R = (Va + Vpe) / Vp
heritability(fit_rep)           # h2 = Va / Vp
permanent_effects(fit_rep)
```

## Common environment

A two-effect model: an additive-genetic effect plus an independent IID
environmental effect (for example a litter or cage), grouped by a separate
column.

```{r}
fit_ce <- hsquared(
  y ~ animal(1 | id, pedigree = ped) + common_env(1 | litter),
  data = dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "two_effect")
  )
)

variance_components(fit_ce)     # animal, common_env, residual
common_env_effects(fit_ce)
```

## Maternal genetic effects

The other two-effect model: a direct additive-genetic effect plus a maternal
genetic effect expressed through the dam. Both carry the pedigree relationship
(the maternal effect uses `A₂ = pedigree A`), and the dams must be animals in the
`animal()` pedigree. This is the independent (uncorrelated) direct–maternal
model; the correlated 2×2 G version is covered at validation scale (opt-in,
`target = "direct_maternal"` — see below).

```{r}
fit_mat <- hsquared(
  y ~ animal(1 | id, pedigree = ped) + maternal_genetic(1 | dam),
  data = dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "two_effect")
  )
)

variance_components(fit_mat)    # animal, maternal_genetic, residual
maternal_effects(fit_mat)
```

## Genomic GREML

A single genomic effect whose relationship is a user-supplied genomic
relationship inverse `Ginv` (with id dimnames) instead of a pedigree. This
estimates the genomic and residual variances, genomic heritability, and genomic
breeding values (GEBVs). You can either supply a precomputed `Ginv`, or pass a
raw marker matrix `M` (rows = individuals) as `genomic(1 | id, markers = M)` and
let the engine build the genomic relationship and its inverse. Single-step can
use either a supplied `Hinv` or the experimental constructed path
`single_step(1 | id, pedigree = ped, markers = M)` with
`target = "single_step_construct"`. External-comparator validation remains
planned.

```{r}
fit_g <- hsquared(
  y ~ genomic(1 | id, Ginv = Ginv),
  data = dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "genomic")
  )
)

variance_components(fit_g)      # genomic, residual
heritability(fit_g)             # genomic h2
breeding_values(fit_g)         # GEBVs
```

## SNP-BLUP / RR-BLUP marker effects

A marker-effect model on a raw marker matrix `genomic(1 | id, markers = M)`. You
may supply the genomic and residual variances (for example from a prior GREML
fit); the engine centres the markers and returns per-marker effects
(`marker_effects()`) together with per-individual genomic breeding values. If you
omit `variance_components`, `hsquared()` instead **estimates** the genomic and
residual variances by REML from the markers (`fit_snp_blup_reml`). Weighted /
Bayesian marker priors and comparator parity are planned.

```{r}
fit_snp <- hsquared(
  y ~ genomic(1 | id, markers = M),
  data = dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(
      target = "snp_blup",
      variance_components = c(sigma_g2 = 1, sigma_e2 = 2)
    )
  )
)

marker_effects(fit_snp)              # one effect per marker
marker_variance_explained(fit_snp)   # descriptive fitted-marker shares
breeding_values(fit_snp)             # per-individual GEBVs
```

## Multivariate Gaussian animal model

The multivariate path uses a `cbind()` response and estimates
trait-by-trait genetic and residual covariance matrices for the animal effect.
It fits on the default call — no `engine` or `target` argument — while the
capability itself stays `partial` and experimental.
Missing response cells are allowed; missing fixed-effect cells are rejected.

```{r}
fit_mv <- hsquared(
  cbind(weight, length) ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE
)

genetic_covariance(fit_mv)
genetic_correlation(fit_mv)
heritability(fit_mv)            # per-trait h2
breeding_values(fit_mv)         # animal-by-trait EBVs
```

The future long-format grammar
`animal(trait | id, pedigree = ped, cov = us())` is planned, not fitted yet. See
the multivariate article for the current claim boundary.

## Binary and count traits (experimental): which heritability to report

<p class="hs-banner" role="note"><strong>Experimental · opt-in · not coverage-calibrated.</strong>
The non-Gaussian animal model fits only through <code>engine = "julia"</code> with
<code>target = "nongaussian"</code>, is limited to an intercept plus one
<code>animal()</code> term, and has not yet been checked against another program
that estimates the same quantity. See <a href="current-limits.html">Can I fit and report this?</a> before reporting a value.</p>

Many traits ecologists and breeders care about are not Gaussian: whether an
individual survived its first winter (0/1), how many of a clutch fledged
(successes out of a known number of trials), how many offspring were sired
(a count). `hsquared()` accepts `binomial(logit)` and `poisson(log)` for these,
and `heritability()` returns **more than one number**, because a
heritability for a non-Gaussian trait depends on the scale you state it on.

Take fledging success in a wild bird population. Each female has a clutch of
between two and seven eggs, and some number of those fledge:

```{r}
# dat: one row per breeding attempt; fledged + failed = clutch size, which varies
fit_fledge <- hsquared(
  cbind(fledged, failed) ~ animal(1 | id, pedigree = ped),
  data = dat,
  family = binomial(),           # logit link
  REML = TRUE,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "nongaussian")   # marginal = "laplace" (default)
  )
)

variance_components(fit_fledge)  # V_A on the latent (logit) scale
heritability(fit_fledge)
```

`heritability()` returns a small table rather than a single value. For a
varying-trials fit like this one it looks like:

```
#>            field                                     label estimate                    undefined_reason
#> 1      h2_latent            latent-scale h2 (conditional)     1.00                                <NA>
#> 2   h2_liability         liability-scale h2 (conditional)     0.20                                <NA>
#> 3 h2_observation observation-scale h2 (no scalar estimand)      NaN varying_trials_no_scalar_estimand
```

(The numbers are illustrative — `V_A = 0.8` on the logit scale gives
$0.8 / (0.8 + \pi^2/3) \approx 0.20$; the shape is exact.)

**Which row do you report? The liability-scale one.** All three rows come
from the same fit on the logit scale; they differ in what goes in the
denominator. Here is what each is, starting with the two that matter:

- **`h2_liability` — report this.** It is the heritability of the *underlying
  liability* to fledge: the additive genetic variance on the logit scale,
  divided by that variance plus the logistic distribution's own variance,
  $\pi^2/3 \approx 3.29$ (the *distribution-specific variance* of Nakagawa and
  Schielzeth 2010). This is the threshold-model convention the field has used
  since Dempster and Lerner (1950), it is the scale on which published
  heritabilities of survival, breeding success and disease traits are usually
  given, and it does not depend on how many trials each individual had. State
  the link when you report it: *"liability-scale $h^2$ (logit link)"*.
- **`h2_observation` — the `NaN` row.** This is the heritability on the *data*
  scale — the 0/1 or proportion you actually recorded. When every individual
  has the same number of trials (a 0/1 survival trait, or a fixed number of
  attempts), this is a single number and the package reports it. **When trial
  sizes vary, it is not a single number**: a female with a clutch of two and
  one with a clutch of seven have different data-scale variances, so no single
  observation-scale $h^2$ exists for the population. `hsquared` therefore
  returns `NaN` with the reason `varying_trials_no_scalar_estimand` rather than
  averaging over clutch sizes and printing something that looks like an
  estimate. **This is not a failed fit.** The liability-scale row is the
  estimate.
- **`h2_latent` — ignore it for now.** It divides the additive variance by the
  *random* latent variance only, with no $\pi^2/3$. In the only model this path
  currently fits — an intercept plus one `animal()` term — the additive variance
  is the sole random component, so this row is always exactly `1`. It is a
  bookkeeping row, not an estimate. The number to keep instead is `V_A` itself,
  from `variance_components()`: it is what every scale conversion, including the
  liability-scale row, starts from.

If you specifically need a data-scale heritability for a varying-trials trait
— for example to predict a response to selection on the observed proportion —
the answer is a per-individual quantity, not a population scalar. Compute it
from the latent-scale `V_A` and intercept with the conversions in
de Villemereuil et al. (2016) (the `QGglmm` package), which integrate over the
link for a stated trial count, or refit on a subset with a common trial count.

Two more things to know about this path. The engine can approximate the
likelihood in two ways — `marginal = "laplace"` (the default) or
`marginal = "variational"` (aliases `"la"`/`"va"`, which maximises a lower
bound on the likelihood) — and their `logLik()` and `AIC()` values are not
comparable with each other, nor with the REML values from the Gaussian fits
above. And Poisson fits report `h2_latent` and a count-scale `h2_observation`
only — there is no liability scale for a count.

*References.* Dempster, E. R. & Lerner, I. M. (1950) Heritability of threshold
characters. *Genetics* 35, 212–236. Nakagawa, S. & Schielzeth, H. (2010)
Repeatability for Gaussian and non-Gaussian data: a practical guide for
biologists. *Biological Reviews* 85, 935–956. de Villemereuil, P., Schielzeth,
H., Nakagawa, S. & Morrissey, M. (2016) General methods for evolutionary
quantitative genetic inference from generalized mixed models. *Genetics* 204,
1281–1294.

## What is fitted, and what is planned

Before extracting an estimate, read [fit diagnostics and plots](visualizing-models.html).
Before reporting one, return to [Can I fit and report this?](current-limits.html).

`formula_status()` and `validation_status()` report, from R, exactly which terms
fit and how each capability is validated.

```{r, eval = TRUE}
library(hsquared)
formula_status()
```

```{r, eval = TRUE}
validation_status()
```

What you may report is listed on [Can I fit and report this?](current-limits.html),
not in `validation_status()`. That function is a developer evidence table;
an absent row is not an absent covered route.
The four opt-in models listed as covered above (`two_effect` common-env leg,
`multi_effect`, `random_regression`, `direct_maternal`) are covered at
validation scale. The remaining opt-in models (`repeatability`, the maternal
two-effect leg, genomic, SNP-BLUP, single-step) remain `partial` until their
`HSquared.jl` gates and external comparators are signed off, and so does the
multivariate `cbind()` model even though it now routes by default.
Structured/factor-analytic multivariate covariance, further non-Gaussian
families (negative binomial, ordinal, zero-inflated, hurdle), unusual-inheritance,
and GPU models remain on the roadmap.
