---
title: "Fitting quantitative-genetic models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting quantitative-genetic models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This article shows the models `hsquared` can fit. Fitting is computed in the
sibling `HSquared.jl` engine, so every example needs a local Julia, the
`JuliaCall` package, and an `HSquared.jl` checkout. The code chunks are not run
when this page is built (the build host has no Julia).

All fits are by REML. ML is not implemented; `REML = FALSE` is rejected on the
fit path.

## The default: a univariate Gaussian animal model

This is the v0.1 model, fitted by the default `hsquared()` call. It estimates the
additive-genetic and residual variances, heritability, and breeding values.

```{r}
library(hsquared)

fit <- hsquared(
  weight ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE
)

variance_components(fit)
heritability(fit)
breeding_values(fit)   # EBVs (aliases: EBV(), BLUP())
summary(fit)
```

The remaining models below are **experimental and opt-in**: they are reached
through `engine = "julia"` with an explicit `target`, and their
`validation_status()` rows are `partial` (they mirror the corresponding
`HSquared.jl` gates, which are still being validated). They are REML-only and are
not yet production- or comparator-validated.

## Repeatability (permanent environment)

A repeated-records model with an additive-genetic effect and a
permanent-environment effect that shares the animal grouping. The additive
(`σ²a`) and permanent-environment (`σ²pe`) variances are only identifiable with
repeated records per individual.

```{r}
fit_rep <- hsquared(
  y ~ animal(1 | id, pedigree = ped) + permanent(1 | id),
  data = repeated_dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "repeatability")
  )
)

variance_components(fit_rep)    # animal, permanent, residual
repeatability(fit_rep)          # R = (Va + Vpe) / Vp
heritability(fit_rep)           # h2 = Va / Vp
permanent_effects(fit_rep)
```

## Common environment

A two-effect model: an additive-genetic effect plus an independent IID
environmental effect (for example a litter or cage), grouped by a separate
column.

```{r}
fit_ce <- hsquared(
  y ~ animal(1 | id, pedigree = ped) + common_env(1 | litter),
  data = dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "two_effect")
  )
)

variance_components(fit_ce)     # animal, common_env, residual
common_env_effects(fit_ce)
```

## Maternal genetic effects

The other two-effect model: a direct additive-genetic effect plus a maternal
genetic effect expressed through the dam. Both carry the pedigree relationship
(the maternal effect uses `A₂ = pedigree A`), and the dams must be animals in the
`animal()` pedigree. This is the independent (uncorrelated) direct–maternal
model; the correlated 2×2 G version is planned.

```{r}
fit_mat <- hsquared(
  y ~ animal(1 | id, pedigree = ped) + maternal_genetic(1 | dam),
  data = dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "two_effect")
  )
)

variance_components(fit_mat)    # animal, maternal_genetic, residual
maternal_effects(fit_mat)
```

## Genomic GREML

A single genomic effect whose relationship is a user-supplied genomic
relationship inverse `Ginv` (with id dimnames) instead of a pedigree. This
estimates the genomic and residual variances, genomic heritability, and genomic
breeding values (GEBVs). You can either supply a precomputed `Ginv`, or pass a
raw marker matrix `M` (rows = individuals) as `genomic(1 | id, markers = M)` and
let the engine build the genomic relationship and its inverse. Single-step
(`Hinv`) and external-comparator validation are planned.

```{r}
fit_g <- hsquared(
  y ~ genomic(1 | id, Ginv = Ginv),
  data = dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "genomic")
  )
)

variance_components(fit_g)      # genomic, residual
heritability(fit_g)             # genomic h2
breeding_values(fit_g)         # GEBVs
```

## SNP-BLUP / RR-BLUP marker effects

A supplied-variance marker-effect model on a raw marker matrix
`genomic(1 | id, markers = M)`. You supply the genomic and residual variances
(for example from a prior GREML fit); the engine centres the markers and returns
per-marker effects (`marker_effects()`) together with per-individual genomic
breeding values. It does not estimate the variance components. REML estimation
of the marker variance, weighted/Bayesian priors, and comparator parity are
planned.

```{r}
fit_snp <- hsquared(
  y ~ genomic(1 | id, markers = M),
  data = dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(
      target = "snp_blup",
      variance_components = c(sigma_g2 = 1, sigma_e2 = 2)
    )
  )
)

marker_effects(fit_snp)              # one effect per marker
marker_variance_explained(fit_snp)   # descriptive fitted-marker shares
breeding_values(fit_snp)             # per-individual GEBVs
```

## Multivariate Gaussian animal model

The opt-in multivariate path uses a `cbind()` response and estimates
trait-by-trait genetic and residual covariance matrices for the animal effect.
Missing response cells are allowed; missing fixed-effect cells are rejected.

```{r}
fit_mv <- hsquared(
  cbind(weight, length) ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "multivariate")
  )
)

genetic_covariance(fit_mv)
genetic_correlation(fit_mv)
heritability(fit_mv)            # per-trait h2
breeding_values(fit_mv)         # animal-by-trait EBVs
```

The future long-format grammar
`animal(trait | id, pedigree = ped, cov = us())` is planned, not fitted yet. See
the multivariate article for the current claim boundary.

## What is fitted, and what is planned

`formula_status()` and `validation_status()` report, from R, exactly which terms
fit and how each capability is validated.

```{r, eval = TRUE}
library(hsquared)
formula_status()
```

```{r, eval = TRUE}
validation_status()
```

Only rows marked `covered` in `validation_status()` may be relied on as working;
the opt-in models above are deliberately `partial` until their `HSquared.jl`
gates and external comparators are signed off. Structured/factor-analytic
multivariate covariance, non-Gaussian/GLLVM, unusual-inheritance, and GPU models
remain on the roadmap.
