---
title: "Formula grammar roadmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formula grammar roadmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This page is a map of the modelling language. It separates syntax that is
parsed today from syntax that is reserved for later phases.

## Parsed today

The current parser contract is intentionally narrow:

```{r, eval = FALSE}
fit <- hsquared(
  y ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE
)
```

The R side validates the formula, data, response, pedigree columns, and animal
IDs. It builds the internal bridge payload for `HSquared.jl`. The default call
still stops at the planned bridge boundary, so this is not general fitted
animal-model support.

## Standard quantitative-genetic extensions

These names are reserved so future models read like the biological question:

```{r, eval = FALSE}
y ~ sex + age +
  animal(1 | id, pedigree = ped) +
  permanent(1 | id) +
  common_env(1 | litter)

y ~ sex + age +
  animal(1 | id, pedigree = ped) +
  maternal_genetic(1 | dam, pedigree = ped) +
  maternal_env(1 | dam) +
  paternal_genetic(1 | sire, pedigree = ped) +
  paternal_env(1 | sire)
```

Current status: these markers are exported but inert. If they appear in an
`hsquared()` formula, the parser errors with planned-not-implemented wording.

## Inheritance and relationship kernels

Later phases should treat inheritance systems as relationship or precision
kernels, not as one-off special cases:

```{r, eval = FALSE}
y ~ animal(1 | id, pedigree = ped) +
  dominance(1 | id, pedigree = ped) +
  epistasis(1 | id, pedigree = ped)

y ~ animal(1 | id, pedigree = ped) +
  cytoplasmic(1 | maternal_line) +
  imprinting(1 | id, pedigree = ped, parent = "maternal")

y ~ relmat(1 | id, K = K_custom)
y ~ precision(1 | id, Q = Q_custom)
```

Current status: syntax reservation only. The engine does not consume these
relationship or precision matrices yet.

## Genomic and marker terms

The genomic vocabulary follows the same rule: readable terms first, fitting
only after the R-Julia contract and validation exist.

```{r, eval = FALSE}
y ~ sex + batch + genomic(1 | id, Ginv = Ginv)

y ~ sex + batch +
  genomic(1 | id, Ginv = Ginv) +
  marker_scan(M, map = marker_map)

y ~ sex + batch + single_step(1 | id, Hinv = Hinv)
```

Current status: `genomic()`, `single_step()`, `markers()`, `marker_scan()`, and
`qtl_scan()` are planned markers. They do not fit genomic, QTL, or eQTL models
yet.

## Multivariate and factor-analytic grammar

The multivariate ladder is planned to connect ordinary G matrices and
GLLVM-style latent genetic structure:

```{r, eval = FALSE}
y ~ trait + trait:sex +
  animal(trait | id, pedigree = ped, cov = us()) +
  residual(trait | unit, cov = us())

y ~ trait + trait:sex +
  animal(trait | id, pedigree = ped, cov = fa(K = 2)) +
  residual(trait | unit, cov = fa(K = 1))
```

Planned covariance meanings:

- `cov = us()`: full unstructured covariance.
- `cov = diag()`: trait-specific variances only.
- `cov = lowrank(K)`: `G = Lambda Lambda'`.
- `cov = fa(K)`: `G = Lambda Lambda' + Psi`.

Current status: not parsed by the v0.1 parser.

## Error rule

Unsupported terms should fail early and plainly. The current expected shape is:

```text
`marker()` is planned, not implemented. The v0.1 parser currently accepts only
`animal(1 | id, pedigree = ped)`.
```

That error is part of the user interface. It tells users the idea is on the
roadmap without pretending that the model has been fitted.

For a compact table of the same status information from R, use:

```{r, eval = FALSE}
formula_status()
```
