---
title: "Formula grammar roadmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formula grammar roadmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This page is a map of the modelling language. It separates syntax that is
parsed today from syntax that is reserved for later phases.

## Parsed today

The current parser contract is intentionally narrow:

```{r, eval = FALSE}
fit <- hsquared(
  y ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE
)
```

When users pass an `hs_data()` bundle with a pedigree component, the same
contract can be written without repeating the pedigree argument:

```{r, eval = FALSE}
bundle <- hs_data(phenotypes = dat, pedigree = ped)

fit <- hsquared(
  y ~ sex + age + animal(1 | id),
  data = bundle,
  family = gaussian(),
  REML = TRUE
)
```

The R side validates the formula, data, response, pedigree columns, and animal
IDs, builds the internal bridge payload, and (by default) fits the v0.1
univariate Gaussian animal model by REML through `HSquared.jl`. Terms below
that say **parsed** are recognized; only some of those also **fit**, and then
only on a named `target =`. Reserved names error as planned, not implemented.

## Standard quantitative-genetic extensions

These names are reserved so future models read like the biological question:

```{r, eval = FALSE}
y ~ sex + age +
  animal(1 | id, pedigree = ped) +
  permanent(1 | id) +
  common_env(1 | litter)

y ~ sex + age +
  animal(1 | id, pedigree = ped) +
  maternal_genetic(1 | dam, pedigree = ped) +
  maternal_env(1 | dam) +
  paternal_genetic(1 | sire, pedigree = ped) +
  paternal_env(1 | sire)
```

Current status: `permanent()`, `common_env()`, and `maternal_genetic()` are
parsed. They fit only with a named `target =` (repeatability, two-effect, or
direct–maternal — see Fitting models). The default `hsquared()` call names
that target and stops. `maternal_env()`, `paternal_genetic()`, and
`paternal_env()` are reserved and error as planned, not implemented.

## Inheritance and relationship kernels

Later phases should treat inheritance systems as relationship or precision
kernels, not as one-off special cases:

```{r, eval = FALSE}
y ~ animal(1 | id, pedigree = ped) +
  dominance(1 | id, pedigree = ped) +
  epistasis(1 | id, pedigree = ped)

y ~ animal(1 | id, pedigree = ped) +
  cytoplasmic(1 | maternal_line) +
  imprinting(1 | id, pedigree = ped, parent = "maternal")

y ~ relmat(1 | id, K = K_custom)
y ~ precision(1 | id, Q = Q_custom)
```

Current status: syntax reservation only. The engine does not consume these
relationship or precision matrices yet.

## Genomic and marker terms

The genomic vocabulary follows the same rule: readable terms first, fitting
only after the R-Julia contract and validation exist.

```{r, eval = FALSE}
y ~ sex + batch + genomic(1 | id, Ginv = Ginv)

y ~ sex + batch +
  genomic(1 | id, Ginv = Ginv) +
  marker_scan(M, map = marker_map)

y ~ sex + batch + single_step(1 | id, Hinv = Hinv)
```

Current status: `markers()`, `marker_scan()`, and `qtl_scan()` are planned
markers and do not fit yet. `genomic()` and `single_step()` now fit through an
opt-in, experimental engine path (`engine = "julia"`, `target = "genomic"` /
`"single_step"` / `"snp_blup"`; REML-only or supplied-variance, not the
default, mirroring a `partial` validation gate).

## Multivariate and factor-analytic grammar

The first multivariate surface is the `cbind()` animal-model path, which routes
on the default call:

```{r, eval = FALSE}
cbind(weight, height) ~ sex + animal(1 | id, pedigree = ped)
```

The longer multivariate ladder is planned to connect ordinary G matrices and
GLLVM-style latent genetic structure:

```{r, eval = FALSE}
y ~ trait + trait:sex +
  animal(trait | id, pedigree = ped, cov = us()) +
  residual(trait | unit, cov = us())

y ~ trait + trait:sex +
  animal(trait | id, pedigree = ped, cov = lowrank(K = 2)) +
  residual(trait | unit, cov = diag())

y ~ trait + trait:sex +
  animal(trait | id, pedigree = ped, cov = fa(K = 2)) +
  residual(trait | unit, cov = fa(K = 1))
```

Planned covariance meanings:

- `cov = us()`: full unstructured covariance.
- `cov = diag()`: trait-specific variances only.
- `cov = lowrank(K)`: `G = Lambda Lambda'`.
- `cov = fa(K)`: `G = Lambda Lambda' + Psi`.

Current status: the `cbind()` multivariate animal-model path is parsed and
routed by default, and fitted experimentally — the routing is default, the
claim is still `partial`. The long-format `animal(trait | id, cov = ...)`
grammar and residual structured covariance grammar are planned, not parsed or
fitted.

## Error rule

Unsupported terms fail early. There are two live shapes.

Reserved names such as `markers()`, `dominance()`, or `maternal_env()` stop
as planned, not implemented, then name the nearest live path:

```text
`markers()` is planned, not implemented. The live marker path is
`genomic(1 | id, markers = )` or `gwas(fit, markers)`. Run
`formula_status()` for the live list of which terms parse and which fit.
```

Terms that already parse but are not on the default path -- `common_env()`,
`permanent()`, `maternal_genetic()`, `rr(...)` -- do not use that planned
stop. They paste the next `hsquared()` call:

```text
`common_env()` is not on the default path.

Closest working call:

  hsquared(
    y ~ animal(1 | id, pedigree = ped) + common_env(1 | litter),
    data = dat,
    control = hs_control(
      engine = "julia",
      engine_control = list(target = "two_effect")
    )
  )

That route is covered for point estimates only (validation-scale,
common-environment leg). Intervals are experimental.
See: current-limits article, formula_status().
```

That error is part of the user interface. It names the closest working call
without pretending the default path already fitted the model.

For a compact table of the same status information from R, use:

```{r, eval = FALSE}
formula_status()
```
