---
title: "Function map and cheat sheet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Function map and cheat sheet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r golden-data}
library(hsquared)

# n = 4 is a syntax demo, not a number for a paper.
ped <- data.frame(
  id   = c("sire", "dam", "off1", "off2"),
  sire = c(NA, NA, "sire", "sire"),
  dam  = c(NA, NA, "dam", "dam")
)
dat <- data.frame(
  id     = c("sire", "dam", "off1", "off2"),
  sex    = c("m", "f", "m", "f"),
  weight = c(42, 38, 40, 37)
)
```

One page for "which function do I call now?". The preview below uses the same
tiny pedigree as the README and Getting started. Every later example either
runs without Julia or is marked as needing the engine.

## The five steps

| Step | You are asking | Call |
|---|---|---|
| **1. Specify** | Is my data in the right shape? Is my model expressible? | `hs_data()`, `data_status()`, `animal()`, `formula_status()` |
| **2. Preview** | What will the engine actually be sent? | `hs_control(engine = "validate")`, `model_spec()` |
| **3. Fit** | Estimate the variance components | `hsquared()` |
| **4. Check** | Did it converge, and may I report it? | `fit_diagnostics()`, `summary()`, [Can I fit and report this?](current-limits.html), `backend_info()` |
| **5. Extract** | Get the quantities for the paper | `heritability()`, `variance_components()`, `breeding_values()`, `fixef()`, `ranef()` |

Steps 1, 2, and 4's status calls need R only. Step 3 and the extractors need a
fitted object, and therefore Julia.

## 1. Specify

```{r specify}
# bundle phenotypes, pedigree, and any genomic/expression/environment metadata
bundle <- hs_data(phenotypes = dat, pedigree = ped)
summary(bundle)
data_status(bundle)   # pedigree coverage, founders, parent links, marker map

# what does the formula grammar parse, reserve, or refuse?
formula_status()
```

`animal(1 | id, pedigree = ped)` is the additive genetic term. With
`data = bundle` you can write `animal(1 | id)` and the bundle pedigree is used.

## 2. Preview — no Julia needed

```{r preview}
hsquared(
  weight ~ sex + animal(1 | id, pedigree = ped),
  data = dat,
  control = hs_control(engine = "validate")
)

model_spec(
  weight ~ sex + animal(1 | id, pedigree = ped),
  data = dat
)
```

This is the cheapest way to find out that a term is a syntax reservation rather
than an implemented model: it tells you before you go looking for Julia.

## 3. Fit — needs the engine

```{r fit, eval = FALSE}
# needs Julia + HSquared.jl; not a silent fallback
fit <- hsquared(
  weight ~ sex + animal(1 | id, pedigree = ped),
  data = dat
)
```

Opt-in experimental targets go through `hs_control(engine = "julia")` with an
`engine_control` target — see
[Can I fit and report this?](current-limits.html) for which ones you may report.

## 4. Check

```{r check, eval = FALSE}
summary(fit)
fit_diagnostics(fit)   # convergence, iterations, at_boundary flag
backend_info()         # which engine answered, and its version
validation_status()    # developer evidence table, not the user list
```

Read `at_boundary` before anything else. A variance pinned at zero is a real
result about your data, not a number to report as an estimate.

## 5. Extract

| You want | Call |
|---|---|
| Heritability | `heritability(fit)` |
| Variance components | `variance_components(fit)` |
| Breeding values (EBVs/BLUPs) | `breeding_values(fit)` |
| Accuracy of those EBVs | `prediction_error_variance(fit)`, `reliability(fit)` |
| Fixed effects | `fixef(fit)`, `coef(fit)` |
| Random effects | `ranef(fit)` |
| Fitted values and residuals | `fitted(fit)`, `residuals(fit)` |
| Model comparison | `logLik(fit)`, `AIC(fit)`, `nobs(fit)` |
| A plot | `plot(fit)`, `autoplot(fit)` |

Route-specific extractors exist for the opt-in targets — `repeatability()`,
`common_env_effects()`, `maternal_effects()`, `genetic_covariance()`,
`marker_effects()`, `g_matrix_geometry()` — and inherit their route's status.

## Uncertainty: read this before using it

```{r intervals, eval = FALSE}
heritability_interval(fit)                    # experimental
variance_component_standard_errors(fit)       # experimental
```

These are asymptotic/delta-method quantities, surfaced when the engine provides
them and labelled experimental. **They are not coverage-calibrated.** No route in
this package currently carries an interval-reporting permission. `confint()` and
`vcov()` are deliberately absent rather than shipped uncalibrated.

## Reserved names that do not fit yet

`qtl_table()`, `gwas_table()`, `eqtl_table()`, `lod_scores()`, and
`marker_variance_explained()` are reserved output names for future marker/QTL
results. They exist so the vocabulary is stable, not because the models fit.
`marker_effects()` is the exception — it is live for the opt-in SNP-BLUP path.

## Where to read next

- [Getting started with hsquared](../hsquared.html) — the guided introduction.
- [Can I fit and report this?](current-limits.html) — reporting permissions per route.
- [Fitting models](fitting-models.html) — the longer fitting guide.
