---
title: "Reading G matrices"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Reading G matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

This article is for the applied moment after a multivariate animal model has
fit: you want to know what the genetic covariance matrix says, what the
residual matrix says, and where the current evidence boundary is.

The current `hsquared` multivariate path is experimental and opt-in. It
estimates an unstructured additive-genetic covariance matrix `G` and residual
covariance matrix `R` for Gaussian animal models. It is not yet production
multi-trait software, and it does not yet provide standard errors, confidence
intervals, selection-response predictions, factor-analytic loadings, or a
general phenotypic covariance extractor.

## Model scale

For the current two-or-more-trait Gaussian animal model,

\[
\mathrm{vec}(\mathbf a) \sim
\mathrm{MVN}\left(\mathbf 0, \mathbf G \otimes \mathbf A\right),
\qquad
\mathrm{vec}(\mathbf e) \sim
\mathrm{MVN}\left(\mathbf 0, \mathbf R \otimes \mathbf I\right).
\]

`A` is the additive relationship matrix from the pedigree. `G` is the
trait-by-trait additive-genetic covariance matrix. `R` is the trait-by-trait
residual covariance matrix. The two matrices are on the scale of the response
traits after the fixed effects in the formula.

```{r}
fit_mv <- hsquared(
  cbind(weight, length) ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "multivariate")
  )
)
```

## The G matrix

Use `G_matrix()` when you want the additive-genetic covariance matrix in the
language of quantitative genetics. It is an alias for the more explicit
`genetic_covariance()` extractor.

```{r}
G_matrix(fit_mv)
genetic_covariance(fit_mv)
```

Read the diagonal as additive-genetic variance for each trait. Larger diagonal
entries mean more additive-genetic variation on that trait's measurement scale.

Read the off-diagonal as additive-genetic covariance between traits. A positive
entry means animals with higher additive-genetic values for one trait tend to
have higher additive-genetic values for the other. A negative entry means the
additive-genetic values tend to move in opposite directions.

For interpretation across traits with different units, use the genetic
correlation:

```{r}
genetic_correlation(fit_mv)
```

The genetic correlation is standardized, so it is often easier to read than the
raw covariance. It is not a causal statement by itself. A correlation tells you
how additive-genetic deviations line up in the fitted model; selection response
still needs the relevant selection gradient and a clearly defined prediction
target.

## The R matrix

Use `R_matrix()` for the residual covariance matrix. It is an alias for
`residual_covariance()`.

```{r}
R_matrix(fit_mv)
residual_covariance(fit_mv)
residual_correlation(fit_mv)
```

`R` collects variation not assigned to the additive animal effect or fixed
effects in the current model. It can include measurement noise, microenvironment,
unmodelled common environment, unmodelled maternal effects, and other residual
sources. Do not read a residual correlation as genetic integration.

## Heritability and EBVs

For the current animal-plus-residual multivariate model, per-trait
heritability is read from the diagonal pieces of `G` and `R`:

\[
h^2_j = \frac{G_{jj}}{G_{jj} + R_{jj}}.
\]

Use the extractor rather than calculating by hand:

```{r}
heritability(fit_mv)
```

Future models with permanent environment, common environment, maternal effects,
or genomic effects will need a model-specific phenotypic denominator. That is
why this slice does not add a general `P_matrix()` extractor yet.

Breeding values are returned by animal and trait:

```{r}
breeding_values(fit_mv)
EBV(fit_mv)
```

In an unbalanced response matrix, an animal can be missing one trait and still
receive breeding values for all traits. The missing trait borrows information
through the fitted genetic covariance and the pedigree relationship.

## What to trust first

Trust invariant summaries first:

- `G_matrix(fit)` / `genetic_covariance(fit)`;
- `R_matrix(fit)` / `residual_covariance(fit)`;
- `genetic_correlation(fit)` and `residual_correlation(fit)`;
- `heritability(fit)`;
- `breeding_values(fit)`;
- `fit_diagnostics(fit)`.

The factor-analytic roadmap will eventually add loadings, specific variance,
latent breeding values, and eigen-G summaries. Those are not just pretty
alternative labels for `G`: loading axes are sign- and rotation-nonunique unless
the model has a validated constraint or rotation policy. Until that policy is
validated, `hsquared` keeps the invariant covariance and correlation matrices
as the stable interpretation layer.

## Current evidence boundary

The current multivariate surface is `partial`.

Covered in the R lane:

- `cbind()` response grammar;
- missing response cells preserved as `NA` and marshalled to Julia `NaN`;
- G/R covariance and correlation extraction;
- `G_matrix()` and `R_matrix()` aliases over those same fields;
- per-trait heritability and cross-trait EBVs;
- a shared deterministic R/Julia fixture;
- an optional `sommer` comparator for the diagonal-residual version of the
  shared fixture.

Still planned:

- a committed `t >= 2` known-truth recovery fixture;
- full external comparator parity for unstructured `G` and full `R`;
- *validated*, coverage-calibrated standard errors and confidence intervals for
  G/R entries, correlations, and heritabilities (experimental, asymptotic SEs are
  already available via `covariance_standard_errors()` for an unstructured fit —
  `V4-MV-REML`, partial: the recovery calibration has not passed, unstructured-only,
  not coverage-calibrated);
- production sparse multivariate fitting;
- `cov = diag()`, `cov = lowrank(K)`, and `cov = fa(K)` R grammar;
- a general `P_matrix()` extractor with a model-specific estimand.

## Background anchors

The reason G matrices matter is classical quantitative genetics: the additive
genetic covariance matrix connects multivariate inheritance, constraint, and
response to selection. Useful anchors are [Lande and Arnold's correlated-trait
selection framework](https://pubmed.ncbi.nlm.nih.gov/28556011/),
[Kirkpatrick, Lofsvold and Bulmer's work on genetic covariance functions for
growth trajectories](https://pubmed.ncbi.nlm.nih.gov/2323560/), and
[Hansen and Houle's evolvability measures for multivariate
characters](https://pubmed.ncbi.nlm.nih.gov/18662244/).

For `hsquared`, those ideas are roadmap direction, not current over-claim. The
current package reports the fitted G and R matrices and keeps future
evolvability, selection-response, and factor-analytic interpretation behind
explicit validation gates.

See also:

- [Multivariate Gaussian animal models](multivariate.html)
- [Model status](model-status.html)
- [Formula grammar roadmap](formula-grammar.html)
