---
title: "Experimental genetic GLLVM for three count traits"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Experimental genetic GLLVM for three count traits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

This experimental route fits three complete Poisson count traits with two
pedigree-linked genetic factors. Use it only when every pedigree animal has
exactly one row with all three responses. The animal ID in `dat` must match
the pedigree; the bridge aligns phenotype rows to pedigree order.

## Model and scale

For animal `i` and trait `t`, the conditional mean is
`exp(beta[t] + u[i,t])`, with `u = F %*% t(Lambda)` and genetic covariance
`G = Lambda %*% t(Lambda)`. The two factors share the pedigree relationship
matrix. The reported trait effects are conditional modes `u` on the **log
rate** scale. They are not factor scores, posterior means, or response-scale
heritabilities.

## Fit and read the result

Replace `dat`, `ped`, and `julia_project` with your complete count table,
pedigree, and local `HSquared.jl` checkout:

```{r}
fit_gllvm <- hsquared(
  cbind(t1, t2, t3) ~ animal(1 | id, pedigree = ped),
  data = dat,
  family = poisson(link = "log"),
  control = hs_control(engine = "julia", engine_control = list(
    target = "genetic_gllvm",
    genetic_structure = "lowrank",
    rank = 2L,
    experimental_gllvm = TRUE,
    julia_project = "/path/to/HSquared.jl",
    iterations = 3000L
  ))
)

fit_diagnostics(fit_gllvm)
G_matrix(fit_gllvm)
genetic_correlation(fit_gllvm)
breeding_values(fit_gllvm)
fixef(fit_gllvm)
```

Check convergence and the diagnostic message before interpreting the
covariance or trait effects. `breeding_values()` returns animal and trait
labels with `scale = "link"`; the underlying fit stores a `q × 3` matrix of
trait genetic modes. Genetic correlations describe the fitted covariance on
the link scale. Raw loadings depend on factor rotation and are withheld.

The diagnostic objective integrates both fixed effects, under a flat
measure, and pedigree genetic effects using a Laplace approximation. It is
not ordinary non-Gaussian maximum likelihood or Gaussian REML, so this fit
does not expose generic `logLik()` or `AIC()` values. The route also does not
report response-scale heritability.

## Current boundary

This is one complete, balanced, three-trait, rank-two, pure low-rank Poisson
cell. It does not fit missing records, repeated records, other response
families, factor-analytic uniqueness, or arbitrary ranks. Automatic rank
selection from the `gllvmTMB` and `GLLVModels.jl` work is a separate follow-on
that needs pedigree-specific validation. A live 12-animal R fit matched a
direct same-input Julia fit for the covariance, correlations, fixed effects,
trait modes, and objective within `1e-8`; this does not establish broad
recovery, calibration, or an external same-objective comparison. The public
covered count remains seven.
