---
title: "Genomics, QTL, and CPU/GPU roadmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genomics, QTL, and CPU/GPU roadmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This page is a roadmap and status map. Some genomic building blocks now exist
as opt-in experimental paths, while production genomics, QTL/eQTL, GLLVM/omics,
and CPU/GPU acceleration remain gated roadmap work.

`hsquared` is designed to grow from a simple animal-model interface into a
coherent modelling language for phenotypes, pedigrees, genotypes, QTL/eQTL,
G matrices, inheritance systems, and high-dimensional traits.

The full technical plan lives in
`docs/design/07-genomics-qtl-gpu-plan.md`. This article is the user-facing
map.

## Design principle

```text
phenotype =
  fixed effects
  + pedigree/genetic effects
  + genomic marker effects
  + maternal/paternal effects
  + environment/common effects
  + latent multivariate factors
  + residual variation
```

The R package should keep this easy to say. The Julia package should provide
the performance-oriented engine, with speed and scale treated as benchmarked
claims rather than slogans.

## Genomic syntax

The basic genomic GREML effect above is the eventual default form. Today it
fits through an opt-in, experimental control (`engine = "julia"`,
`target = "genomic"`; a raw `markers = M` matrix and a SNP-BLUP
`target = "snp_blup"` path are also available) — see the "Fitting models"
article. The default `engine = "fit"` call shown here is the planned future
surface.

```{r, eval = FALSE}
fit <- hsquared(
  y ~ sex + batch + genomic(1 | id, Ginv = Ginv),
  data = pheno,
  family = gaussian()
)
```

Marker and scan syntax is planned later:

```{r, eval = FALSE}
fit <- hsquared(
  y ~ sex + age +
    genomic(1 | id, Ginv = Ginv) +
    marker_scan(M, map = marker_map, leave_one_chr_out = TRUE),
  data = pheno,
  genotypes = geno,
  family = gaussian()
)
```

## Planned backend strategy

CPU is the trusted default. GPU is an accelerator path that must be
benchmarkable and numerically checked.

Planned Julia backends:

- CPU and threaded CPU;
- CUDA for NVIDIA GPUs;
- Metal for Mac/Apple Silicon testing;
- AMDGPU for ROCm systems;
- oneAPI for Intel accelerators;
- portable kernels where appropriate.

## Planned QTL and eQTL path

The intended scan grammar is:

```{r, eval = FALSE}
fit <- hsquared(
  y ~ sex + age +
    genomic(1 | id, Ginv = Ginv) +
    marker_scan(M, map = marker_map, leave_one_chr_out = TRUE),
  data = pheno,
  genotypes = geno,
  family = gaussian()
)

gwas_table(fit)
plot_manhattan(fit)
plot_qq(fit)
```

For expression and high-dimensional omics, the long-run plan is to combine
marker scans, sample factors, and genomic relatedness:

```{r, eval = FALSE}
fit <- hsquared(
  expr_matrix ~ batch + sex +
    marker_scan(M, map = marker_map) +
    sample_factors(K = 5) +
    genomic(1 | id, Ginv = Ginv),
  data = expr_data,
  genotypes = geno,
  family = gaussian()
)
```

The marker-scan, multi-term, and high-dimensional multivariate examples above
are syntax targets only and do not fit models today. The single genomic,
single-step, SNP-BLUP, and simple `cbind()` multivariate animal-model paths fit
only opt-in and experimentally (see "Fitting models").

## Planned multivariate and GLLVM path

Classical multivariate animal models and GLLVM-style models should sit on one
ladder:

```{r, eval = FALSE}
fit <- hsquared(
  y ~ trait + trait:sex +
    animal(trait | id, pedigree = ped, cov = fa(K = 2)) +
    residual(trait | unit, cov = fa(K = 1)),
  data = long_dat,
  family = gaussian()
)
```

The same idea should later extend to response matrices, count families,
ordination, latent breeding values, and high-dimensional G matrices.

## Evidence gate

The package will not claim ASReml-level performance, GPU speedup, production
genomic support, QTL/eQTL support, or GLLVM support until code, tests, docs,
and validation evidence exist. Current genomic and marker-scan paths are
experimental, opt-in, and partial.
