---
title: "Genomic prediction"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Genomic prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

This article shows the current opt-in genomic prediction surface in `hsquared`.
It is for users who already have a genomic relationship inverse, a marker
matrix, or a single-step relationship inverse and want to understand which path
is live today.

The current genomic paths are experimental, REML-only unless stated otherwise,
and Julia-backed. They are not the default `hsquared()` fit. They require
`control = hs_control(engine = "julia", engine_control = list(...))`, a local
Julia, `JuliaCall`, and a sibling `HSquared.jl` checkout.

## Three current paths

The current R surface has three distinct genomic paths.

| Path | Formula term | Target | Variances |
| --- | --- | --- | --- |
| GBLUP / genomic GREML from a supplied inverse | `genomic(1 | id, Ginv = Ginv)` | `target = "genomic"` | estimated by the Julia AI-REML path |
| GBLUP / genomic GREML from raw markers | `genomic(1 | id, markers = M)` | `target = "genomic"` | estimated after the engine builds and regularizes G |
| SNP-BLUP / RR-BLUP marker effects | `genomic(1 | id, markers = M)` | `target = "snp_blup"` | supplied by the user |

There is also a single-step surface:

| Path | Formula term | Target | Variances |
| --- | --- | --- | --- |
| supplied-H inverse single-step | `single_step(1 | id, Hinv = Hinv)` | `target = "single_step"` | estimated by the Julia AI-REML path |

Building `Hinv` from a pedigree and genomic relationship is planned. APY,
low-rank marker solvers, weighted genomic relationships, Bayesian marker
models, GWAS, QTL, and eQTL scans are also planned.

## Supplied Ginv

Use this path when you already have a genomic relationship inverse whose row and
column names match the IDs in the phenotype data.

```{r}
fit_g <- hsquared(
  y ~ batch + genomic(1 | id, Ginv = Ginv),
  data = pheno,
  family = gaussian(),
  REML = TRUE,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "genomic")
  )
)

variance_components(fit_g)
heritability(fit_g)
breeding_values(fit_g)
fit_diagnostics(fit_g)
```

Read the random effect as a genomic breeding value: \(\mathbf g \sim
N(0, \mathbf G \sigma_g^2)\), with `Ginv` supplied on the inverse scale.

The parser checks that every observed ID is present in the `Ginv` dimnames.
The current path estimates the genomic and residual variance components by the
Julia AI-REML engine. It is still partial because broad comparator parity and
production-scale genomic validation are not complete.

## Marker-built G

Use this path when you have an individual-by-marker dosage matrix `M`, with
rownames matching the phenotype IDs.

```{r}
fit_marker_g <- hsquared(
  y ~ batch + genomic(1 | id, markers = M),
  data = pheno,
  family = gaussian(),
  REML = TRUE,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "genomic")
  )
)

variance_components(fit_marker_g)
heritability(fit_marker_g)
breeding_values(fit_marker_g)
```

The engine centers the marker matrix, builds a genomic relationship matrix, and
uses a regularized inverse for the REML fit. This is convenient for small and
validation-scale examples. It is not yet a production genotype pipeline:
PLINK/VCF readers, imputation hooks, scaling/blending choices, APY, and large
on-disk marker workflows remain future work.

## SNP-BLUP at supplied variances

Use this path when you want marker effects at variance components you already
trust, for example from a prior genomic GREML fit. This is a supplied-variance
solve, not variance-component estimation.

```{r}
fit_snp <- hsquared(
  y ~ batch + genomic(1 | id, markers = M),
  data = pheno,
  family = gaussian(),
  REML = TRUE,
  control = hs_control(
    engine = "julia",
    engine_control = list(
      target = "snp_blup",
      variance_components = c(sigma_g2 = 1.0, sigma_e2 = 2.0)
    )
  )
)

marker_effects(fit_snp)
breeding_values(fit_snp)
heritability(fit_snp)
fit_diagnostics(fit_snp)
```

`marker_effects(fit_snp)` returns one effect per marker column. The genomic
breeding values are the marker-derived predictions for each genotyped
individual. The fitted object records `variance_components_source = "supplied"`
because this target uses the supplied `sigma_g2` and `sigma_e2`.

## Supplied Hinv single-step

Use this path only when you already have a single-step inverse relationship
matrix `Hinv` with dimnames matching the phenotype IDs.

```{r}
fit_h <- hsquared(
  y ~ batch + single_step(1 | id, Hinv = Hinv),
  data = pheno,
  family = gaussian(),
  REML = TRUE,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "single_step")
  )
)

variance_components(fit_h)
heritability(fit_h)
breeding_values(fit_h)
```

This is not yet automatic single-step HBLUP construction. The package does not
currently build `Hinv` from pedigree and genotype inputs. That construction is
planned as a separate validated relationship-matrix slice.

## Output checklist

For the current genomic paths, start with:

```{r}
fit_diagnostics(fit_g)
variance_components(fit_g)
heritability(fit_g)
breeding_values(fit_g)
```

For SNP-BLUP, also use:

```{r}
marker_effects(fit_snp)
```

The QTL/GWAS/eQTL output names are reserved, but they are not fitted by the
current genomic paths:

```{r}
qtl_table(fit_g)    # planned unless a future QTL target produced it
gwas_table(fit_g)   # planned
eqtl_table(fit_g)   # planned
```

## Current evidence boundary

The current genomic and single-step rows are `partial`.

Covered in the R lane:

- parser and payload checks for supplied `Ginv`, marker matrices, supplied
  `Hinv`, and SNP-BLUP variance inputs;
- skip-guarded live bridge tests for genomic GREML, marker-built G, supplied
  `Hinv`, and SNP-BLUP when Julia and the sibling engine are available;
- extractor checks for variance components, genomic heritability, genomic
  breeding values, marker effects, and provenance diagnostics.

Still planned:

- production relationship-matrix construction and scaling/blending controls;
- building `Hinv` from pedigree plus genotype data;
- APY and low-rank large-marker workflows;
- PLINK/VCF/BCF readers and on-disk marker storage;
- genomic comparator parity with sommer, BLUPF90-family tools, JWAS, or other
  agreed benchmarks;
- marker scans, QTL, GWAS, and eQTL;
- uncertainty intervals and reliability for production genomic predictions.

## Background anchors

The current syntax follows standard genomic prediction ideas: marker-based
prediction of total genetic value, genomic relationship matrices for GBLUP, and
single-step blending of pedigree and genomic information. Useful anchors are
[Meuwissen, Hayes and Goddard (2001)](https://pubmed.ncbi.nlm.nih.gov/11290733/),
[VanRaden (2008)](https://pubmed.ncbi.nlm.nih.gov/18946147/), and
[Aguilar, Misztal, Johnson, Legarra, Tsuruta and Lawlor
(2010)](https://pubmed.ncbi.nlm.nih.gov/20105546/).

For `hsquared`, those papers motivate the roadmap. The current package exposes
small, opt-in genomic building blocks and keeps broad production genomic claims
behind explicit validation gates.

See also:

- [Fitting models](fitting-models.html)
- [Genomics, QTL, and CPU/GPU roadmap](genomics-gpu-roadmap.html)
- [Model status](model-status.html)
