---
title: "A worked animal-model analysis (gryphon)"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<p class="hs-banner" role="note"><strong>Intervals: asymptotic / experimental · not coverage-calibrated.</strong>
This is a teaching walk of the default univariate animal-model path on gryphon. Whiskers, when shown, are not coverage-calibrated 95% CIs for publication.</p>

This article walks one univariate animal model end to end on a real teaching dataset — fit,
heritability with its experimental uncertainty, breeding values and accuracy, and a diagnostic
plot — so you can see the whole `hsquared` extractor surface in one place. It complements the
broad [Fitting models](fitting-models.html) tour and the
[sommer / pedigreemm benchmark](benchmark-comparators.html) (which checks the same fit against
external packages).

Two honesty notes first:

- **Fitting needs a local engine.** The default `hsquared()` fit marshals to a local Julia +
  `HSquared.jl` + `JuliaCall`. The code below is shown but not executed at build time; the
  heritability value quoted is the published/recovered estimate (see the benchmark article for the
  executed agreement).
- **gryphon uses a supplied relationship matrix.** The open Wilson et al. 2010
  gryphon pedigree has no ancestral loops (parents precede offspring; no
  self-parents). This article still uses the supplied `A_gryphon` matrix as the
  signed-off recovery path. Most of your own datasets will use the ordinary
  `animal(1 | id, pedigree = ped)` path shown in [Fitting models](fitting-models.html).

## The data and question

The gryphon birth-weight dataset (`enhancer::DT_gryphon` / `A_gryphon`, from Wilson et al. 2010,
*An ecologist's guide to the animal model*) is the canonical teaching example. The question is the
classic one: **how heritable is birth weight?** — i.e. what fraction of phenotypic variance is
additive genetic, `h² = σ²a / (σ²a + σ²e)`.

```{r data}
library(hsquared)
utils::data("DT_gryphon", package = "enhancer")
str(DT_gryphon[, c("ANIMAL", "BWT")])
```

## Fit the animal model

The univariate Gaussian animal model is `BWT ~ 1 + animal`, by REML. With a raw pedigree you would
write `hsquared(BWT ~ animal(1 | id, pedigree = ped), data = dat)`; for gryphon the relationship
information is supplied via the engine path.

```{r fit}
# Requires a local Julia + HSquared.jl + JuliaCall.
fit <- hsquared(BWT ~ animal(1 | id, pedigree = ped), data = dat)
fit
```

The first fit in a session takes about 30 s while Julia starts and compiles.
Later fits take under a second for a few hundred animals. The call prints no
progress, so the first one can look frozen.

## Read the result

The published REML estimates are `σ²a = 3.3954`, `σ²e = 3.8286`, giving **`h² = 0.470`** — birth
weight is moderately heritable. `summary()` reports the point estimates and, when a local engine
provides them, the **experimental** uncertainty surfaces (asymptotic, REML-only — see
[Validation evidence](validation-evidence.html)):

```{r summary}
summary(fit)
heritability(fit)          # point estimate (≈ 0.470)
heritability_interval(fit) # experimental logit delta-method CI, if available
heritability_standard_error(fit)
variance_component_standard_errors(fit)
```

The interval and standard errors are reported as point-estimate-plus-bounds only; they are not
coverage-calibrated and should not be read as validated inference.

## Breeding values and accuracy

```{r ebv}
ebv <- breeding_values(fit)   # one predicted additive genetic value per animal
head(ebv)
accuracy(fit)                 # sqrt(reliability), when reliability is present
```

## A diagnostic look

```{r plot}
plot(fit, type = "variance")   # components, with experimental +/- 1.96 SE whiskers if present
plot(fit, type = "residuals")  # residuals vs fitted
```

## Interpreting `h² = 0.47`

A heritability near one-half means roughly half the variation in birth weight among these animals
is additive genetic — there is appreciable raw material for a response to selection, and the
predicted breeding values rank animals by genetic merit for the trait. As always, `h²` is a
property of *this population in this environment*, not a fixed constant of the trait.

## Honest boundaries

- The fit requires a local engine; this article uses the supplied `A_gryphon`
  relationship matrix as the signed-off recovery path. The open gryphon
  pedigree has no ancestral loops. Your own analyses will typically use
  `animal(1 | id, pedigree = ped)`.
- The CI/SE surfaces are **experimental** (engine row `V1-HERIT-CI`, `partial`): asymptotic,
  REML-only, unreliable at small `n`. They are reported, not validated. The
  Wald SE and logit-delta CI are not valid when additive variance is near 0.
  The likelihood-ratio test of `V_A = 0` uses a 50:50 mixture of `chi^2_0`
  and `chi^2_1` (Self and Liang 1987). R `anova()` does not run it; Julia
  `nested_lrt(..., df = 1, boundary_df = 1)` already does.
- gryphon is teaching/simulated data; the published numbers are an external anchor, not a
  known-truth simulation. For bias/coverage and external-package agreement see
  [Validation evidence](validation-evidence.html) and the
  [benchmark](benchmark-comparators.html).

## See also

- [Visualizing models](visualizing-models.html) — diagnose a fitted result before
  reporting it.
- [Can I fit and report this?](current-limits.html) — the reporting permission
  for this route.
- [Fitting models](fitting-models.html) — the broad tour, including the ordinary raw-pedigree path
  and the opt-in models.
- [Benchmark: hsquared vs sommer and pedigreemm](benchmark-comparators.html) — the same gryphon fit
  checked against external REML fitters.
