---
title: "Inheritance systems roadmap"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Inheritance systems roadmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

<p class="hs-banner" role="note"><strong>Experimental · opt-in · not the default fit.</strong>
Inheritance-system terms beyond diploid additive animal are roadmap / planned or partial — this article does not activate them.</p>

This article shows how `hsquared` is expected to grow beyond the ordinary
diploid additive animal model.

The core design rule is simple:

```text
inheritance system -> relationship matrix K or precision matrix Q
model term         -> design matrix Z
engine target      -> variance component and BLUP/EBV calculation
```

That rule keeps animal breeding, plant breeding, evolutionary ecology, and
genomic work on one modelling ladder. It also keeps the public API from turning
into a bag of special cases.

## What works today

The default fit is the v0.1 additive animal model:

```{r fixture}
ped <- data.frame(
  id   = c("s1", "s2", "d1", "d2", "o1", "o2", "o3", "o4"),
  sire = c(NA, NA, NA, NA, "s1", "s1", "s2", "s2"),
  dam  = c(NA, NA, NA, NA, "d1", "d2", "d1", "d2")
)
pheno <- data.frame(
  id  = ped$id,
  sex = c("m", "m", "f", "f", "m", "f", "m", "f"),
  age = c(4, 5, 4, 5, 1, 1, 1, 1),
  y   = c(72, 75, 65, 68, 31, 33, 35, 30)
)
```

```{r}
fit <- hsquared(
  y ~ sex + age + animal(1 | id, pedigree = ped),
  data = pheno,
  family = gaussian(),
  REML = TRUE
)
```

The following standard extensions fit only through opt-in experimental engine
targets:

```{r}
# permanent environment / repeatability
y ~ animal(1 | id, pedigree = ped) + permanent(1 | id)

# IID common environment
y ~ animal(1 | id, pedigree = ped) + common_env(1 | litter)

# independent direct additive + maternal genetic effect
y ~ animal(1 | id, pedigree = ped) + maternal_genetic(1 | dam)
```

Those are not unusual-inheritance models. They are useful stepping stones:
extra random effects, pedigree-linked second effects, and clear extractor
boundaries.

## Planned named effects

Some future effects already have reserved formula markers:

```{r}
y ~ animal(1 | id, pedigree = ped) +
  maternal_env(1 | dam)

y ~ animal(1 | id, pedigree = ped) +
  paternal_genetic(1 | sire, pedigree = ped) +
  paternal_env(1 | sire)

y ~ animal(1 | id, pedigree = ped) +
  cytoplasmic(1 | maternal_line)

y ~ animal(1 | id, pedigree = ped) +
  imprinting(1 | id, pedigree = ped, parent = "maternal")

y ~ animal(1 | id, pedigree = ped) +
  dominance(1 | id, pedigree = ped) +
  epistasis(1 | id, pedigree = ped)
```

These examples do not fit yet. Today the parser rejects them as planned, not
implemented.

The supplied-relationship primaries `relmat(1 | id, K = K_custom)` and
`precision(1 | id, Q = Q_custom)` are the exception: they **do** fit through the
opt-in `engine = "julia"` path (the same supplied-relationship-inverse engine as
`genomic()`). They are REML-only, experimental, and **not** covered — supply a
dense positive-definite `K` (or its inverse `Q`), and the matrix is carried as
provenance, not estimated.

## Supplying your own relationship matrix

Three spellings take a user matrix, and they are not interchangeable:

| Term | Argument | What you pass | Inverse |
| --- | --- | --- | --- |
| `relmat(1 \| id, K = K)` | `K` | the relationship / covariance itself | R inverts it (`Kinv = solve(K)`) |
| `precision(1 \| id, Q = Q)` | `Q` (or `Kinv`) | the precision | used as given |
| `genomic(1 \| id, Ginv = Ginv)` | `Ginv` | the genomic relationship **inverse** | used as given |

`K`, `Q`, and `Ginv` must be square, symmetric, finite, and positive definite,
with unique non-missing row and column names equal to the ids. Rows may be in
any order; they are matched by name, not by position. The fitted coefficient
from `relmat()` is `Va / (Va + Ve)` on the supplied kernel, not a pedigree
heritability unless `K` is the pedigree `A`.

A monozygotic (MZ) twin, clone, or repeated-genotype kernel is singular by
construction: identical individuals have identical rows, so the smallest
eigenvalue is zero and the parser stops with `` `K` must be positive definite ``.
This package does not accept that singular `K` and does not invent a new
decomposition. The AE workaround is a small ridge, `K + 1e-6 * diag(n)`.
Variance `Va` and the REML log-likelihood match the unridged kernel; the
printed residual is `E - 1e-6 * Va`, so printed `h2 = Va / (Va + Ve)` is
slightly inflated unless you add `1e-6 * Va` back to the denominator. A
larger `eps` moves residual and printed `h2` further while leaving `Va`
unchanged. An ACE twin model (`relmat()` plus a pair / common-environment
term) is not available from R; collapsing each MZ pair to one genotype is
the design alternative. This is not an ACE estimator.

```{r}
# One MZ pair (K = 1) and one DZ pair (K = 0.5). Unrelated between pairs.
twin_dat <- data.frame(
  id = c("mz1", "mz2", "dz1", "dz2"),
  zyg = c("MZ", "MZ", "DZ", "DZ"),
  y = c(1.10, 1.02, 0.40, 0.55)
)
K <- diag(nrow(twin_dat))
rownames(K) <- colnames(K) <- twin_dat$id
K["mz1", "mz2"] <- K["mz2", "mz1"] <- 1
K["dz1", "dz2"] <- K["dz2", "dz1"] <- 0.5
# The raw MZ kernel is singular. Ridge it before fitting.
K_fit <- K + 1e-6 * diag(nrow(K))
fit_k <- hsquared(
  y ~ relmat(1 | id, K = K_fit),
  data = twin_dat,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "relmat")
  )
)
variance_components(fit_k)
heritability(fit_k) # Va / (Va + Ve) on the ridged kernel
```

## Planned inheritance systems

Other systems are roadmap examples rather than exported syntax:

```{r}
# planned future shapes, not implemented
animal(1 | id, pedigree = ped, inheritance = selfing(rate = s))
animal(1 | id, pedigree = ped, inheritance = clonal())
animal(1 | id, pedigree = ped, inheritance = haplodiploid())
animal(1 | id, pedigree = ped, inheritance = polyploid(ploidy = 4))
animal(1 | id, pedigree = ped, inheritance = cytoplasmic())
```

These names are useful design placeholders, but the package does not currently
export `selfing()`, `clonal()`, `haplodiploid()`, `polyploid()`, or an
`inheritance =` grammar for `animal()`.

## Why kernels first

Most inheritance systems become mixed-model terms only after the relationship
or precision structure is clear.

| System | Kernel question | Public risk if rushed |
| --- | --- | --- |
| selfing / partial selfing | How does inbreeding change additive covariance through the pedigree? | A standard diploid A matrix can give the wrong expectation. |
| clonal / asexual | Which records share genotypes, ramets, environments, or clone lines? | Clone identity can be confounded with common environment. |
| haplodiploid | How do sex-specific inheritance and relatedness enter A or Q? | Parent-offspring and sib covariances are not the ordinary diploid ones. |
| polyploid | What ploidy, double reduction, and dosage assumptions define relatedness? | Diploid shortcuts can misstate plant-breeding relationships. |
| cytoplasmic | What is the maternal-line transmission structure? | It can be mistaken for maternal environment. |
| imprinting | Which parent-of-origin effect is modelled? | Maternal and paternal allele-origin effects can be conflated. |
| dominance / epistasis | Which non-additive relationship matrix is used? | Additive h2 can be over-interpreted as total genetic variance. |
| custom K / Q | Who constructed the matrix, and on what scale? | Scale, ordering, and positive-definiteness mistakes become silent model errors. |

The engine should prefer sparse precision matrices when possible. If a system
cannot supply a well-defined `K` or `Q`, the formula should stay planned.

## Validation gates

An inheritance system should not move beyond `planned` until all of these are
true:

- mathematical definition of the estimand and relationship/precision matrix;
- tiny deterministic pedigree or genotype example;
- ID ordering and missing-parent behaviour pinned by tests;
- positive-definite or positive-semidefinite checks where relevant;
- comparator check against an established package or a published worked
  example when one exists;
- recovery simulation for the variance component and predicted effects;
- extractor and summary wording that separates additive, non-additive,
  maternal, paternal, cytoplasmic, and environmental interpretations.

For dominance and epistasis, `nadiv` is a natural comparison target. For
diploid/autopolyploid A/G/H relationship construction, `AGHmatrix` is a natural
comparison target. For breeding-program simulations, `XSim.jl` is a natural
future partner.

## User-facing rule

The common path should stay easy:

```{r}
y ~ fixed + animal(1 | id, pedigree = ped)
```

Specialist inheritance should be explicit:

```{r}
# planned
y ~ fixed +
  animal(1 | id, pedigree = ped) +
  cytoplasmic(1 | maternal_line)
```

The user should never need to guess whether "maternal" means maternal genetic,
maternal environmental, cytoplasmic, parent-of-origin, or dam identity. Names
should stay longer until the biology is unambiguous.

## Background anchors

Useful software anchors include
[nadiv](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2012.00213.x)
for non-additive genetic relatedness matrices and
[AGHmatrix](https://pubmed.ncbi.nlm.nih.gov/37471595/) for diploid and
autopolyploid A/G/H relationship construction. The package-level roadmap also
tracks custom relationship and precision matrices because many plant, animal,
and ecological designs will arrive as a known kernel before they arrive as a
named inheritance helper.

For now, these sources motivate the roadmap. They are not evidence of current
support for selfing, clonal, haplodiploid, polyploid, cytoplasmic, imprinting,
dominance, epistasis, or custom-kernel models.

See also:

- [Formula grammar roadmap](formula-grammar.html)
- [Fitting models](fitting-models.html)
- [Model status](model-status.html)
