---
title: "Multivariate Gaussian animal models"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Multivariate Gaussian animal models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE
)
```

Use the multivariate path when the same animals have two or more quantitative
traits and the question is about genetic covariance: which traits share
additive-genetic variation, what are the genetic correlations, and what are the
trait-specific heritabilities?

This path is experimental and opt-in. It is useful for tiny validation-scale
Gaussian examples, and the validation ledger now records cold-start recovery,
one full-unstructured `sommer` REML comparator, a published supplied-covariance
Mrode anchor, and a `MCMCglmm` Bayesian agreement probe. It is still not an
ASReml-style production multi-trait claim and is not promoted beyond `partial`.
The examples are not run when this page is built because they require a local
Julia, `JuliaCall`, and a sibling `HSquared.jl` checkout.

## Model

For two traits measured on animal records, the fitted model is

\[
\mathrm{vec}(\mathbf a) \sim
\mathrm{MVN}\left(\mathbf 0, \mathbf G_0 \otimes \mathbf A\right),
\qquad
\mathrm{vec}(\mathbf e) \sim
\mathrm{MVN}\left(\mathbf 0, \mathbf R_0 \otimes \mathbf I\right).
\]

Here `A` is the pedigree additive relationship matrix, `G0` is the trait-by-trait
additive-genetic covariance matrix, and `R0` is the trait-by-trait residual
covariance matrix. `hsquared` sends an NA-preserving response matrix to
`HSquared.jl`; missing response cells are masked by the engine. Missing
fixed-effect values are still rejected before fitting.

## Current syntax

The current live grammar uses a `cbind()` response and the ordinary animal-model
random intercept:

```{r}
fit_mv <- hsquared(
  cbind(weight, length) ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "multivariate")
  )
)
```

Read this as one additive animal effect shared across traits. The trait
covariances live in `G0`; the residual covariances live in `R0`.

## Missing trait cells

Trait cells may be missing in the `cbind()` response. For example, this is a
valid data shape for the multivariate target:

```{r}
dat <- data.frame(
  id = c("sire", "dam", "calf1", "calf2"),
  sex = c("m", "f", "f", "m"),
  age = c(4, 4, 1, 1),
  weight = c(72, 65, 31, 35),
  length = c(110, NA, 72, 76)
)

ped <- data.frame(
  id = c("sire", "dam", "calf1", "calf2"),
  sire = c(NA, NA, "sire", "sire"),
  dam = c(NA, NA, "dam", "dam")
)

fit_mv <- hsquared(
  cbind(weight, length) ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE,
  control = hs_control(
    engine = "julia",
    engine_control = list(target = "multivariate")
  )
)
```

The missing `length` value is allowed because it is a response cell. A missing
`sex`, `age`, or `id` value would be rejected because those variables define the
fixed-effect and random-effect design matrices.

## Extracting results

The multivariate fit returns the same kind of object as the univariate fit, plus
trait covariance matrices:

```{r}
genetic_covariance(fit_mv)
residual_covariance(fit_mv)

G_matrix(fit_mv)
R_matrix(fit_mv)

genetic_correlation(fit_mv)
residual_correlation(fit_mv)

heritability(fit_mv)
breeding_values(fit_mv)
fit_diagnostics(fit_mv)
```

`heritability(fit_mv)` is per trait. `breeding_values(fit_mv)` returns EBVs by
animal and trait. `fit_diagnostics(fit_mv)` is the first place to look before
interpreting likelihood-based summaries; `logLik()` and `AIC()` are intentionally
blocked for non-converged multivariate fits.

`covariance_standard_errors(fit_mv)` returns **experimental**, asymptotic
delta-method standard errors for the G/R covariance and correlation matrices and
per-trait h², when a local engine provides them. It mirrors `V4-MV-REML`
(`partial`): unstructured-only, REML-only, not coverage-calibrated. The strict
per-seed recovery gate is still a non-pass (7/12 seeds), but the 12-seed
bias/MCSE study (twin `HSquared.jl#78`/`#79`) shows no detectable bias
(`|bias| ≤ 2·MCSE` for all six covariance parameters) — reported, not validated.

## Initial values

If you supply starting values, use a named list. The names are part of the bridge
contract:

```{r}
fit_mv <- hsquared(
  cbind(weight, length) ~ sex + age + animal(1 | id, pedigree = ped),
  data = dat,
  family = gaussian(),
  REML = TRUE,
  control = hs_control(
    engine = "julia",
    engine_control = list(
      target = "multivariate",
      initial = list(
        G0 = diag(1, 2),
        R0 = diag(1, 2)
      )
    )
  )
)
```

A bare two-element list is rejected because it is too easy to swap genetic and
residual covariance starts by accident.

## Planned structured grammar

The long-format structured covariance grammar is planned, not implemented:

```{r}
# Planned, not fitted yet:
y ~ trait + sex + age +
  animal(trait | id, pedigree = ped, cov = us())

# Planned, not fitted yet:
y ~ trait + sex + age +
  animal(trait | id, pedigree = ped, cov = fa(K = 2))
```

Use the `cbind()` path above for the current opt-in multivariate model. Expert
users can select the current rotation-free diagonal-G bridge through
`engine_control = list(genetic_structure = "diagonal")`, but the formula grammar
shown here remains reserved for later unstructured, diagonal, low-rank, and
factor-analytic covariance models.

## Claim boundary

This page describes a partial capability:

- fitted only through `engine = "julia"` and `target = "multivariate"`;
- Gaussian and REML-only;
- animal-model-only;
- dense validation-scale;
- missing response cells allowed, missing fixed-effect cells rejected;
- no `cov = us()`, `cov = diag()`, `cov = lowrank(K)`, or `cov = fa(K)` grammar
  yet;
- R-lane evidence includes a 100-replicate cold-start recovery study, one
  full-unstructured `sommer` REML comparator, a published Mrode Example 5.1
  supplied-covariance BLUP/MME anchor, and a `MCMCglmm` Bayesian agreement
  probe;
- covered promotion still needs the recovery gate accepted or broadened plus a
  second independent same-estimand comparator beyond `sommer`;
- `MCMCglmm` is Bayesian/MCMC agreement evidence, not same-estimand REML
  parity.

Use `validation_status()` and the model-status article before treating any
multivariate result as more than an experimental fit.
