---
title: "Visualizing an animal model"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Visualizing an animal model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.4,
  dpi = 120
)
```

`hsquared` ships a small `ggplot2` visualization layer in the style of the
`brms`/`bayesplot` ecosystem (and consistent with the sister packages `drmTMB`
and `gllvmTMB`). Every figure is produced by an `autoplot()` method, returns a
`ggplot` object you can theme and extend, and is **uncertainty-first**: where the
fit carries the experimental standard errors, reliabilities, or prediction error
variances, the figure shows them — clearly labelled experimental and asymptotic.

In your own analysis, `fit` is the object returned by `hsquared()`:

```{r, eval = FALSE}
fit <- hsquared(
  weight ~ sex + animal(1 | id, pedigree = ped),
  data = dat, family = gaussian(), REML = TRUE
)
autoplot(fit, "variance")
```

The figures below are rendered from **small illustrative fit objects** so the
article is self-contained (no engine required to build the page); the
`autoplot()` calls are exactly what you would run on a real fit.

```{r setup, include = FALSE}
library(hsquared)
library(ggplot2)
set.seed(1)

# --- illustrative univariate fit ---
n <- 80
bv <- sort(stats::rnorm(n, 0, 0.7))
fit <- structure(
  list(result = list(
    variance_components = data.frame(
      component = c("animal", "residual"), estimate = c(0.45, 0.92)
    ),
    variance_component_se = data.frame(
      component = c("animal", "residual"), se = c(0.13, 0.10)
    ),
    heritability = data.frame(term = "trait", estimate = 0.33),
    heritability_se = 0.07,
    breeding_values = data.frame(id = as.character(seq_len(n)), value = bv),
    prediction_error_variance = data.frame(
      id = as.character(seq_len(n)),
      value = stats::runif(n, 0.18, 0.30)
    )
  )),
  class = "hsquared_fit"
)

# --- illustrative multivariate fit (genetic correlation + covariance) ---
g <- matrix(
  c(1, 0.42, -0.18, 0.42, 1, 0.27, -0.18, 0.27, 1), 3, 3,
  dimnames = list(c("mass", "length", "horn"), c("mass", "length", "horn"))
)
sds <- c(1.1, 0.9, 0.6)
g_cov <- outer(sds, sds) * g
fit_mv <- structure(
  list(result = list(genetic_correlation = g, genetic_covariance = g_cov)),
  class = "hsquared_fit"
)

# --- illustrative random-regression fit (order-2 K_g + residual variance) ---
fit_rr <- structure(
  list(
    spec = list(target = "random_regression"),
    result = list(
      coefficient_covariance = matrix(c(0.30, 0.045, 0.045, 0.07), 2, 2),
      residual_variance = 0.55,
      random_regression = list(
        covariate = "age", order = 2L, lower = 1, upper = 5
      )
    )
  ),
  class = "hsquared_fit"
)

# --- illustrative marker scan ---
m <- 400
pv <- stats::runif(m)
pv[c(120, 250)] <- c(2e-4, 8e-5)
gw <- structure(
  data.frame(
    marker = paste0("snp", seq_len(m)), effect = stats::rnorm(m),
    se = stats::runif(m, 0.1, 0.3), z = stats::rnorm(m),
    chisq = stats::rchisq(m, 1), p_value = pv,
    bonferroni_p = pmin(1, pv * m), bh_qvalue = pv, lod = -log10(pv) / 2
  ),
  class = c("hs_gwas", "data.frame")
)
```

## Variance components and heritability

`autoplot(fit, "variance")` draws a horizontal forest of the variance components
and per-trait `h²`, with approximate 95% intervals (`± 1.96 · SE`) when the fit
carries the experimental standard errors.

```{r}
autoplot(fit, "variance")
```

## Breeding values

`autoplot(fit, "breeding_values")` draws the sorted estimated breeding values as
a caterpillar, with `± 1.96 · √PEV` bands when prediction error variances are
available (faceted by trait for multivariate fits).

```{r}
autoplot(fit, "breeding_values")
```

## Genetic correlation (G)

`autoplot(fit, "g_matrix")` draws the genetic-correlation matrix of an opt-in
multivariate fit. It is **rotation-invariant** — correlations do not depend on
the factor rotation — so it is well defined for any multivariate fit; raw factor
loadings are never plotted (the ratified cross-lane convention).

```{r, fig.width = 5.2, fig.height = 4.4}
autoplot(fit_mv, "g_matrix")
```

The heatmap also flags off-diagonal cells that involve a low-`h²` trait (where the
genetic correlation is imprecise); pass `low_h2 =` to set the threshold.

## Genetic eigenstructure (G geometry)

`autoplot(fit_mv, "g_geometry")` draws a scree of the **rotation-invariant** genetic
eigenstructure — the eigenvalues of `G` (variance per genetic axis) with the percent
variance explained. Axis directions and loadings are never drawn: they are
rotation-arbitrary and span-ambiguous under repeated eigenvalues.

```{r, fig.width = 5.6, fig.height = 4}
autoplot(fit_mv, "g_geometry")
```

## Marker scan (Manhattan)

`autoplot()` on the result of `gwas(fit, markers)` draws a Manhattan plot. The
p-values are **nominal Wald p-values and are not genome-wide calibrated** — the
figure carries that banner, and the dashed line is a plain Bonferroni reference,
not a validated genome-wide threshold. HSquared.jl PR #134 adds a fixed-panel
calibration smoke harness on the engine side, but this plot still does not carry
an activated R significance threshold.

```{r}
autoplot(gw)
```

## Marker scan (QQ)

`autoplot(gw, "qq")` draws the QQ plot of the scan: observed versus expected
`−log10(p)` with a `y = x` null reference and the genomic-inflation `lambda_GC` as a
**diagnostic** annotation. As with the Manhattan, the p-values are nominal Wald
p-values and are not genome-wide calibrated.

```{r, fig.width = 5.2, fig.height = 4.4}
autoplot(gw, "qq")
```

## Reaction norms (random regression)

For a random-regression fit, `autoplot(fit, "reaction_norm")` draws the
genetic-variance and heritability trajectories across the covariate. The `h²(t)`
trajectory is experimental: with a homogeneous residual and no permanent-environment
term it can overstate heritability for repeated-records designs.

```{r}
autoplot(fit_rr, "reaction_norm")
```

`autoplot(fit, "rr_eigenfunctions")` draws the rotation-invariant eigenfunctions
`ψ_j(t)` of the coefficient covariance `K_g` as covariate functions, faceted by axis
and labelled with the percent genetic variance. Eigenfunction signs are arbitrary
and the curves are span-ambiguous under repeated eigenvalues, so do not over-read
their direction.

```{r}
autoplot(fit_rr, "rr_eigenfunctions")
```

`autoplot(fit, "rr_surface")` draws the genetic covariance surface
`S(s, t) = φ(s)′ K_g φ(t)` over the covariate grid as a heatmap; pass
`correlation = TRUE` for the genetic-correlation surface.

```{r, fig.width = 5.2, fig.height = 4.4}
autoplot(fit_rr, "rr_surface")
```

## Known-truth recovery studies

`hs_recovery_forest()` visualizes a known-truth recovery study (such as
`data-raw/multivariate-recovery-study.R`): each target's bias with its
`± 2 · MCSE` interval. Intervals that cover zero indicate **no detectable bias**.

```{r}
recovery <- data.frame(
  target = c("G0[1,1]", "G0[2,1]", "G0[2,2]", "R0[1,1]", "R0[2,1]",
             "R0[2,2]", "rg", "h2[1]", "h2[2]"),
  bias = c(-0.00395, 0.00621, -0.01459, 0.01171, -0.01223, -0.00106,
           0.01652, -0.00756, -0.00676),
  mcse = c(0.02221, 0.01397, 0.01726, 0.01387, 0.01164, 0.01199,
           0.01553, 0.00817, 0.00697)
)
hs_recovery_forest(recovery)
```

## Theming

Every figure returns a `ggplot` object, so you can restyle it with the usual
`ggplot2` grammar, and `theme_hsquared()` is exported for reuse:

```{r, eval = FALSE}
autoplot(fit, "variance") +
  ggplot2::labs(title = "My trait") +
  theme_hsquared(base_size = 14)
```
